---
title: "Derivation of tissue-specific proteomic aging signatures from Ding et al. (2025)"
subtitle: "Age-associated proteins, proteome-transcriptome decoupling, and proteomic aging-clock features"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
    toc_float: true
---

```{r global, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
```

# Paths, packages, and compatibility checks

```{r paths}
current_input <- knitr::current_input(
  dir = TRUE
)

if (
  length(current_input) == 1L &&
    !is.na(current_input) &&
    nzchar(current_input)
) {
  RMD_DIR <- dirname(
    normalizePath(
      current_input,
      winslash = "/",
      mustWork = TRUE
    )
  )
} else {
  RMD_DIR <- normalizePath(
    getwd(),
    winslash = "/",
    mustWork = TRUE
  )
}

REFERENCE_DIR <- file.path(
  RMD_DIR,
  "reference_data"
)

OUTDIR <- file.path(
  RMD_DIR,
  "outputs"
)

S2FILE <- Sys.getenv(
  "DING2025_S2_FILE",
  unset = ""
)

S3FILE <- Sys.getenv(
  "DING2025_S3_FILE",
  unset = ""
)

if (!nzchar(S2FILE)) {
  stop(
    paste0(
      "DING2025_S2_FILE is not set. ",
      "Set it to the Ding 2025 mmc2 supplementary workbook."
    ),
    call. = FALSE
  )
}

if (!nzchar(S3FILE)) {
  stop(
    paste0(
      "DING2025_S3_FILE is not set. ",
      "Set it to the Ding 2025 mmc3 supplementary workbook."
    ),
    call. = FALSE
  )
}

stopifnot(
  file.exists(S2FILE),
  file.exists(S3FILE)
)

dir.create(
  REFERENCE_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  OUTDIR,
  recursive = TRUE,
  showWarnings = FALSE
)
```

```{r packages}
library(dplyr)
library(purrr)
library(readr)
library(readxl)
library(stringr)
library(tibble)
library(tidyr)

user_lib <- Sys.getenv(
  "SIGREPO_R_LIB",
  unset = ""
)

if (nzchar(user_lib)) {
  stopifnot(
    dir.exists(user_lib)
  )

  .libPaths(
    c(
      user_lib,
      .libPaths()
    )
  )
}

library(OmicSignature)

find_codex_reference_root <- function(start = getwd()) {
  current <- normalizePath(
    start,
    winslash = "/",
    mustWork = TRUE
  )

  repeat {
    candidate <- file.path(
      current,
      "codex_reference"
    )

    if (dir.exists(candidate)) {
      return(
        normalizePath(
          candidate,
          winslash = "/",
          mustWork = TRUE
        )
      )
    }

    parent <- dirname(current)

    if (identical(parent, current)) {
      stop(
        "Could not locate the codex_reference directory.",
        call. = FALSE
      )
    }

    current <- parent
  }
}

CODEX_REFERENCE_ROOT <- find_codex_reference_root(
  RMD_DIR
)

OMIC_HELPER <- file.path(
  CODEX_REFERENCE_ROOT,
  "helpers",
  "omicsignature_compat.R"
)

stopifnot(
  file.exists(OMIC_HELPER)
)

source(
  OMIC_HELPER,
  local = TRUE
)

OMIC_API <- sigrepo_assert_compatible_api()
print(OMIC_API)
```

# Preflight interpretation

## Study and source tables

- Study: Ding et al., *Cell*, 2025.
- PMID: `40713952`.
- DOI: `10.1016/j.cell.2025.06.047`.
- Table S2, worksheet `DEPs`: tissue-specific age-associated proteins with
  study-provided UniProt accessions, Spearman correlations, raw P values, and
  source direction labels.
- Table S2, worksheet `Decoupling proteins`: tissue-specific
  proteome-transcriptome decoupling proteins reported by gene symbol.
- Table S3, worksheet `Clock importance proteins`: tissue-specific proteins
  with permutation-importance values from the proteomic aging-clock models.

## Biologically justified objects

1. Thirteen tissue-specific age-associated protein signatures.
2. Six tissue-specific proteome-transcriptome decoupling signatures.
3. Thirteen tissue-specific proteomic aging-clock feature signatures.
4. Three collections, one for each biologically related signature family.

The supplied worksheets contain selected result tables rather than complete
parent differential-analysis tables. Therefore, `difexp = NULL` for every
object.

## Study-supported score and membership rules

### Age-associated proteins

The study defines age-dependent proteins using the exact criterion
`P_value < 0.05`. Positive Spearman correlations indicate increasing protein
abundance with age; negative correlations indicate decreasing abundance with
age. The signed Spearman correlation is retained as `score`.

### Proteome-transcriptome decoupling

The study defines decoupling proteins as proteins with positive protein-mRNA
correlation in young samples and a decrease in correlation during aging. The
reported change is `R_Old - R_Young`. Because this signature represents the
magnitude of a non-directional loss of coupling, `score = abs(Diff)` and the
workflow verifies that `Diff < 0` and agrees numerically with
`R_Old - R_Young`.

### Proteomic aging-clock features

The study identifies model features using permutation importance. Figure S5A
uses the exact criterion `Importance > 1`; this workflow applies that same
strict operator and retains `Importance` as the positive score.

## Covariates

- The age-associated protein analysis uses unadjusted Spearman correlation
  between chronological age and protein abundance.
- The decoupling analysis uses group-specific Spearman correlations and does
  not report adjustment covariates.
- The aging-clock analysis uses chronological age as the prediction target,
  not as an adjustment covariate.

Accordingly, `covariates = "none"` is used for all three source families.

## Identifier and probe-ID strategy

- Study-provided UniProt accessions in the DEP worksheet are cleaned and used
  directly as `feature_name`. No claim is made that these source accessions
  were revalidated against current UniProt status.
- Multi-accession DEP cells are expanded into separate rows while preserving
  the original unsplit cell and source statistics in the mapping audit.
- Symbol-only decoupling and clock worksheets use a reviewed, frozen
  source-label-to-UniProt cache.
- Exact `Protein` labels that also occur in the Table S2 `DEPs` worksheet use
  the study-provided UniProt accession from that worksheet. The exact
  study-specific crosswalk takes precedence over generic HGNC mappings because
  the source distinguishes entries such as `CUX1` versus `CUX1.1`.
- Source labels absent from the `DEPs` worksheet use the HGNC `2025-04-01`
  archive `uniprot_ids` field. Direct approved symbols take precedence, with
  supported unambiguous previous-symbol rescue where necessary.
- Candidate HGNC one-to-many mappings are not expanded when the study's own
  exact protein-label crosswalk identifies one measured UniProt entry.
- No genuine assay probe identifier is supplied in these worksheets. The
  compatibility helper standardizes each final signature once and assigns the
  repository-compatible `feature_1`, `feature_2`, ... probe IDs.

# Cleaning helpers

```{r cleaning-helpers}
clean_uniprot <- function(x) {
  x <- as.character(x)

  x <- gsub(
    "[[:space:]\u00A0\u2007\u202F\uFEFF]+",
    "",
    x,
    perl = TRUE
  )

  x <- gsub(
    '^"|"$',
    "",
    x
  )

  toupper(x)
}

is_uniprot_accession <- function(x) {
  grepl(
    paste0(
      "^(?:",
      "[OPQ][0-9][A-Z0-9]{3}[0-9]",
      "|",
      "[A-NR-Z][0-9](?:[A-Z][A-Z0-9]{2}[0-9]){1,2}",
      ")(?:-[0-9]+)?$"
    ),
    as.character(x),
    perl = TRUE
  )
}

clean_symbol <- function(x) {
  x <- trimws(
    as.character(x)
  )

  toupper(x)
}

nonempty <- function(x) {
  !is.na(x) & nzchar(x)
}
```

# Import and preserve the source tables

```{r import-source-tables}
s2_dec_tbl <- readxl::read_excel(
  path = S2FILE,
  sheet = "Decoupling proteins",
  skip = 3,
  col_names = c(
    "Tissue",
    "Protein",
    "R_Young",
    "R_Old",
    "Diff"
  )
) |>
  dplyr::mutate(
    source_file = basename(S2FILE),
    source_sheet = "Decoupling proteins",
    source_row = dplyr::row_number() + 3L,
    Tissue = trimws(as.character(Tissue)),
    Protein = clean_symbol(Protein),
    R_Young = as.numeric(R_Young),
    R_Old = as.numeric(R_Old),
    Diff = as.numeric(Diff)
  ) |>
  dplyr::filter(
    nonempty(Tissue),
    nonempty(Protein)
  )

s2_dep_tbl <- readxl::read_excel(
  path = S2FILE,
  sheet = "DEPs",
  skip = 3,
  col_names = c(
    "Tissue",
    "Protein",
    "Uniprot_ID",
    "Description",
    "Spearman_r",
    "P_value",
    "Type"
  )
) |>
  dplyr::mutate(
    source_file = basename(S2FILE),
    source_sheet = "DEPs",
    source_row = dplyr::row_number() + 3L,
    Tissue = trimws(as.character(Tissue)),
    Protein = clean_symbol(Protein),
    original_uniprot_cell = trimws(
      as.character(Uniprot_ID)
    ),
    Description = as.character(Description),
    Spearman_r = as.numeric(Spearman_r),
    P_value = as.numeric(P_value),
    Type = trimws(as.character(Type)),
    Uniprot_ID = stringr::str_replace_all(
      original_uniprot_cell,
      "\\s*(?:;|\\||,)\\s*",
      ";"
    )
  ) |>
  dplyr::filter(
    nonempty(Tissue),
    nonempty(Protein),
    nonempty(original_uniprot_cell)
  ) |>
  tidyr::separate_rows(
    Uniprot_ID,
    sep = ";"
  ) |>
  dplyr::mutate(
    Uniprot_ID = clean_uniprot(Uniprot_ID)
  ) |>
  dplyr::filter(
    nonempty(Uniprot_ID)
  ) |>
  dplyr::group_by(
    source_file,
    source_sheet,
    source_row
  ) |>
  dplyr::mutate(
    source_accession_index = dplyr::row_number(),
    original_feature_name = original_uniprot_cell,
    feature_name = Uniprot_ID,
    primary_mapping_key = original_uniprot_cell,
    rescue_mapping_key = NA_character_,
    mapping_source = "study-provided UniProt accession",
    mapping_multiplicity = dplyr::n_distinct(
      feature_name
    ),
    match_status = dplyr::case_when(
      mapping_multiplicity == 1L ~ "mapped",
      mapping_multiplicity > 1L ~ "one_to_many",
      TRUE ~ "unmatched"
    )
  ) |>
  dplyr::ungroup()

s3_clock_tbl <- readxl::read_excel(
  path = S3FILE,
  sheet = "Clock importance proteins",
  skip = 3,
  col_names = c(
    "Tissue",
    "Protein",
    "Importance_LCL_5",
    "Importance",
    "Importance_UCL_95",
    "Permutation_Error"
  )
) |>
  dplyr::mutate(
    source_file = basename(S3FILE),
    source_sheet = "Clock importance proteins",
    source_row = dplyr::row_number() + 3L,
    Tissue = trimws(as.character(Tissue)),
    Protein = clean_symbol(Protein),
    Importance_LCL_5 = as.numeric(Importance_LCL_5),
    Importance = as.numeric(Importance),
    Importance_UCL_95 = as.numeric(Importance_UCL_95),
    Permutation_Error = as.numeric(Permutation_Error)
  ) |>
  dplyr::filter(
    nonempty(Tissue),
    nonempty(Protein)
  )

cat("S2 DEP tissues:\n")
print(table(s2_dep_tbl$Tissue))

cat("\nS2 decoupling tissues:\n")
print(table(s2_dec_tbl$Tissue))

cat("\nS3 clock tissues:\n")
print(table(s3_clock_tbl$Tissue))

EXPECTED_DEP_SOURCE_ROWS <- c(
  "Aorta" = 3862L,
  "Spleen" = 2679L,
  "Adrenal" = 2262L,
  "Pancreas" = 2262L,
  "Intestine" = 1464L,
  "Lymph" = 1171L,
  "Heart" = 990L,
  "Skin" = 977L,
  "Muscle" = 907L,
  "Liver" = 804L,
  "Adipose" = 760L,
  "Lung" = 684L,
  "Plasma" = 338L
)

EXPECTED_DECOUPLING_SOURCE_ROWS <- c(
  "Spleen" = 251L,
  "Muscle" = 99L,
  "Lymph" = 347L,
  "Adipose" = 464L,
  "Liver" = 1154L,
  "Adrenal" = 1265L
)

EXPECTED_CLOCK_SOURCE_ROWS <- c(
  "Adipose" = 196L,
  "Adrenal" = 551L,
  "Aorta" = 275L,
  "Heart" = 579L,
  "Intestine" = 325L,
  "Liver" = 278L,
  "Lung" = 253L,
  "Lymph" = 299L,
  "Muscle" = 199L,
  "Pancreas" = 515L,
  "Plasma" = 39L,
  "Skin" = 56L,
  "Spleen" = 342L
)

actual_named_counts <- function(x) {
  counts <- table(x)
  stats::setNames(
    as.integer(counts),
    names(counts)
  )
}

stopifnot(
  nrow(s2_dep_tbl) == 19160L,
  nrow(s2_dec_tbl) == 3580L,
  nrow(s3_clock_tbl) == 3907L,
  identical(
    actual_named_counts(s2_dep_tbl$Tissue)[
      names(EXPECTED_DEP_SOURCE_ROWS)
    ],
    EXPECTED_DEP_SOURCE_ROWS
  ),
  identical(
    actual_named_counts(s2_dec_tbl$Tissue)[
      names(EXPECTED_DECOUPLING_SOURCE_ROWS)
    ],
    EXPECTED_DECOUPLING_SOURCE_ROWS
  ),
  identical(
    actual_named_counts(s3_clock_tbl$Tissue)[
      names(EXPECTED_CLOCK_SOURCE_ROWS)
    ],
    EXPECTED_CLOCK_SOURCE_ROWS
  )
)

message(
  "Expected Ding source-table row counts passed."
)
```

# Validate source-defined tissues and study rules

```{r source-validation}
dep_tissues <- c(
  "Aorta",
  "Spleen",
  "Adrenal",
  "Pancreas",
  "Intestine",
  "Lymph",
  "Heart",
  "Skin",
  "Muscle",
  "Liver",
  "Adipose",
  "Lung",
  "Plasma"
)

dec_tissues <- c(
  "Spleen",
  "Muscle",
  "Lymph",
  "Adipose",
  "Liver",
  "Adrenal"
)

clock_tissues <- c(
  "Adipose",
  "Adrenal",
  "Aorta",
  "Heart",
  "Intestine",
  "Liver",
  "Lung",
  "Lymph",
  "Muscle",
  "Pancreas",
  "Plasma",
  "Skin",
  "Spleen"
)

stopifnot(
  setequal(
    unique(s2_dep_tbl$Tissue),
    dep_tissues
  ),
  setequal(
    unique(s2_dec_tbl$Tissue),
    dec_tissues
  ),
  setequal(
    unique(s3_clock_tbl$Tissue),
    clock_tissues
  )
)

stopifnot(
  all(
    is.finite(s2_dep_tbl$P_value)
  ),
  all(
    s2_dep_tbl$P_value < 0.05
  ),
  all(
    is.finite(s2_dep_tbl$Spearman_r)
  ),
  all(
    s2_dep_tbl$Spearman_r != 0
  ),
  all(
    s2_dep_tbl$Type[
      s2_dep_tbl$Spearman_r > 0
    ] == "Upregulated"
  ),
  all(
    s2_dep_tbl$Type[
      s2_dep_tbl$Spearman_r < 0
    ] == "Downregulated"
  )
)

decoupling_difference_error <- abs(
  s2_dec_tbl$Diff -
    (
      s2_dec_tbl$R_Old -
        s2_dec_tbl$R_Young
    )
)

stopifnot(
  all(
    is.finite(s2_dec_tbl$R_Young)
  ),
  all(
    is.finite(s2_dec_tbl$R_Old)
  ),
  all(
    is.finite(s2_dec_tbl$Diff)
  ),
  all(
    s2_dec_tbl$R_Young > 0
  ),
  all(
    s2_dec_tbl$Diff < 0
  ),
  all(
    decoupling_difference_error < 1e-6
  )
)

CLOCK_IMPORTANCE_CUTOFF <- 1

stopifnot(
  all(
    is.finite(s3_clock_tbl$Importance)
  ),
  any(
    s3_clock_tbl$Importance >
      CLOCK_IMPORTANCE_CUTOFF
  )
)

message(
  "Source tissue sets and study-supported selection rules passed."
)
```

# Load the frozen reviewed source-label-to-UniProt mapping cache

```{r symbol-mapping}
MAPPING_CACHE <- file.path(
  REFERENCE_DIR,
  "Ding2025_HGNC_to_UniProt_2025-04-01.rds"
)

if (!file.exists(MAPPING_CACHE)) {
  stop(
    paste0(
      "The reviewed Ding 2025 source-label-to-UniProt mapping cache is missing: ",
      MAPPING_CACHE,
      ". Run finalize_ding2025_reviewed_mapping_cache.R after generating the ",
      "static candidate, then render this Rmd."
    ),
    call. = FALSE
  )
}

symbol_map <- readRDS(
  MAPPING_CACHE
)

required_mapping_columns <- c(
  "source_symbol",
  "feature_name",
  "resolved_hgnc_symbol",
  "mapping_method",
  "mapping_source"
)

stopifnot(
  is.data.frame(symbol_map),
  all(
    required_mapping_columns %in%
      names(symbol_map)
  )
)

symbol_map <- symbol_map |>
  dplyr::transmute(
    source_symbol = clean_symbol(source_symbol),
    feature_name = clean_uniprot(feature_name),
    resolved_hgnc_symbol = clean_symbol(
      resolved_hgnc_symbol
    ),
    mapping_method = as.character(
      mapping_method
    ),
    mapping_source = as.character(
      mapping_source
    )
  ) |>
  dplyr::filter(
    nonempty(source_symbol),
    nonempty(feature_name),
    nonempty(resolved_hgnc_symbol),
    nonempty(mapping_method),
    nonempty(mapping_source)
  ) |>
  dplyr::distinct()

all_symbols_to_map <- sort(
  unique(
    c(
      s2_dec_tbl$Protein,
      s3_clock_tbl$Protein
    )
  )
)

stopifnot(
  !anyNA(symbol_map),
  !anyDuplicated(symbol_map),
  nrow(symbol_map) == 5370L,
  dplyr::n_distinct(
    symbol_map$source_symbol
  ) == 5370L,
  !anyDuplicated(
    symbol_map$source_symbol
  ),
  !anyDuplicated(
    symbol_map[
      c(
        "source_symbol",
        "feature_name"
      )
    ]
  ),
  setequal(
    symbol_map$source_symbol,
    all_symbols_to_map
  ),
  sum(
    symbol_map$mapping_method ==
      "study_provided_exact_label"
  ) == 4700L,
  sum(
    symbol_map$mapping_method ==
      "approved_symbol"
  ) == 667L,
  sum(
    symbol_map$mapping_method ==
      "previous_symbol"
  ) == 3L,
  all(
    is_uniprot_accession(
      symbol_map$feature_name
    )
  )
)

message(
  paste0(
    "Loaded the reviewed Ding 2025 mapping cache: ",
    nrow(symbol_map),
    " one-to-one source-label mappings."
  )
)

apply_symbol_mapping <- function(df) {
  df |>
    dplyr::mutate(
      source_symbol = clean_symbol(Protein)
    ) |>
    dplyr::left_join(
      symbol_map,
      by = "source_symbol",
      relationship = "many-to-many"
    ) |>
    dplyr::group_by(
      source_file,
      source_sheet,
      source_row,
      Tissue,
      source_symbol
    ) |>
    dplyr::mutate(
      original_feature_name = source_symbol,
      primary_mapping_key = source_symbol,
      rescue_mapping_key = dplyr::if_else(
        !is.na(mapping_method) &
          mapping_method %in%
            c(
              "previous_symbol",
              "alias_symbol"
            ),
        resolved_hgnc_symbol,
        NA_character_
      ),
      mapping_source = dplyr::coalesce(
        mapping_source,
        paste0(
          "HGNC 2025-04-01 archive; ",
          "no UniProt mapping after supported rescue"
        )
      ),
      mapping_multiplicity = dplyr::n_distinct(
        feature_name[
          nonempty(feature_name)
        ]
      ),
      match_status = dplyr::case_when(
        mapping_multiplicity == 0L ~ "unmatched",
        mapping_multiplicity == 1L ~ "mapped",
        mapping_multiplicity > 1L ~ "one_to_many"
      )
    ) |>
    dplyr::ungroup()
}

s2_dec_mapped <- apply_symbol_mapping(
  s2_dec_tbl
)

s3_clock_mapped <- apply_symbol_mapping(
  s3_clock_tbl
)
```

# Mapping QC

```{r mapping-qc}
cat("\nS2 decoupling mapping status:\n")
print(
  table(
    s2_dec_mapped$match_status,
    useNA = "ifany"
  )
)

cat("\nS3 clock mapping status:\n")
print(
  table(
    s3_clock_mapped$match_status,
    useNA = "ifany"
  )
)

dec_unmatched <- s2_dec_mapped |>
  dplyr::filter(
    match_status == "unmatched"
  ) |>
  dplyr::distinct(
    source_symbol
  )

clock_unmatched <- s3_clock_mapped |>
  dplyr::filter(
    match_status == "unmatched"
  ) |>
  dplyr::distinct(
    source_symbol
  )

dec_one_to_many <- s2_dec_mapped |>
  dplyr::filter(
    match_status == "one_to_many"
  ) |>
  dplyr::distinct(
    source_symbol,
    feature_name,
    resolved_hgnc_symbol,
    mapping_method
  )

clock_one_to_many <- s3_clock_mapped |>
  dplyr::filter(
    match_status == "one_to_many"
  ) |>
  dplyr::distinct(
    source_symbol,
    feature_name,
    resolved_hgnc_symbol,
    mapping_method
  )

cat("\nS2 decoupling unmatched source symbols:\n")
print(dec_unmatched)

cat("\nS3 clock unmatched source symbols:\n")
print(clock_unmatched)

cat("\nS2 decoupling one-to-many mappings:\n")
print(dec_one_to_many)

cat("\nS3 clock one-to-many mappings:\n")
print(clock_one_to_many)

mapping_summary <- tibble::tibble(
  source_family = c(
    "Decoupling",
    "Clock"
  ),
  source_rows = c(
    nrow(s2_dec_tbl),
    nrow(s3_clock_tbl)
  ),
  mapped_source_symbols = c(
    dplyr::n_distinct(
      s2_dec_mapped$source_symbol[
        s2_dec_mapped$match_status %in%
          c("mapped", "one_to_many")
      ]
    ),
    dplyr::n_distinct(
      s3_clock_mapped$source_symbol[
        s3_clock_mapped$match_status %in%
          c("mapped", "one_to_many")
      ]
    )
  ),
  mapped_rows = c(
    sum(
      s2_dec_mapped$match_status %in%
        c("mapped", "one_to_many")
    ),
    sum(
      s3_clock_mapped$match_status %in%
        c("mapped", "one_to_many")
    )
  ),
  unresolved_source_symbols = c(
    nrow(dec_unmatched),
    nrow(clock_unmatched)
  )
)

print(mapping_summary)

expected_mapping_summary <- tibble::tibble(
  source_family = c(
    "Decoupling",
    "Clock"
  ),
  source_rows = c(
    3580L,
    3907L
  ),
  mapped_source_symbols = c(
    3069L,
    3211L
  ),
  mapped_rows = c(
    3580L,
    3907L
  ),
  unresolved_source_symbols = c(
    0L,
    0L
  )
)

stopifnot(
  identical(
    mapping_summary,
    expected_mapping_summary
  ),
  nrow(dec_unmatched) == 0L,
  nrow(clock_unmatched) == 0L,
  nrow(dec_one_to_many) == 0L,
  nrow(clock_one_to_many) == 0L
)

message(
  "Reviewed Ding mapping coverage passed with no unresolved or one-to-many source labels."
)
```

```{r optional-audit-write}
do_write_audit <- FALSE

if (do_write_audit) {
  readr::write_csv(
    s2_dep_tbl,
    file.path(
      OUTDIR,
      "Ding2025_DEP_mapping_audit.csv"
    )
  )

  readr::write_csv(
    s2_dec_mapped,
    file.path(
      OUTDIR,
      "Ding2025_decoupling_mapping_audit.csv"
    )
  )

  readr::write_csv(
    s3_clock_mapped,
    file.path(
      OUTDIR,
      "Ding2025_clock_mapping_audit.csv"
    )
  )
}
```

# Metadata constants

```{r metadata-constants}
sample_type_map <- c(
  "Aorta" = "aorta",
  "Spleen" = "spleen",
  "Adrenal" = "adrenal gland",
  "Pancreas" = "pancreas",
  "Intestine" = "intestine",
  "Lymph" = "lymph node",
  "Heart" = "heart",
  "Skin" = "skin",
  "Muscle" = "skeletal muscle",
  "Liver" = "liver",
  "Adipose" = "white adipose tissue",
  "Lung" = "lung",
  "Plasma" = "blood plasma"
)

name_token_map <- c(
  "Aorta" = "Aorta",
  "Spleen" = "Spleen",
  "Adrenal" = "AdrenalGland",
  "Pancreas" = "Pancreas",
  "Intestine" = "Intestine",
  "Lymph" = "LymphNode",
  "Heart" = "Heart",
  "Skin" = "Skin",
  "Muscle" = "SkeletalMuscle",
  "Liver" = "Liver",
  "Adipose" = "WhiteAdiposeTissue",
  "Lung" = "Lung",
  "Plasma" = "BloodPlasma"
)

stopifnot(
  setequal(
    names(sample_type_map),
    dep_tissues
  ),
  setequal(
    names(name_token_map),
    dep_tissues
  )
)

purrr::walk(
  unname(sample_type_map),
  sigrepo_assert_sample_type
)

AUTHOR <- "Yingjie Ding et al."
PMID_STR <- "40713952"
YEAR_NUM <- 2025
PLATFORM_STR <- "proteomics by mass spectrometry"
ASSAY_STR <- "proteomics"
COVARIATES_STR <- "none"

sigrepo_assert_platform(
  PLATFORM_STR
)
```

# Shared object validation

```{r validation-helpers}
validate_signature_table <- function(
    sig,
    expected_levels,
    context,
    require_exact_level_order = TRUE
) {
  level_check <- if (
    isTRUE(require_exact_level_order)
  ) {
    identical(
      levels(sig$group_label),
      expected_levels
    )
  } else {
    all(
      levels(sig$group_label) %in%
        expected_levels
    )
  }

  stopifnot(
    is.data.frame(sig),
    all(
      c(
        "probe_id",
        "feature_name",
        "score",
        "group_label"
      ) %in% names(sig)
    ),
    nrow(sig) > 0L,
    is.character(sig$probe_id),
    all(
      nonempty(sig$probe_id)
    ),
    !anyDuplicated(sig$probe_id),
    identical(
      sig$probe_id,
      paste0(
        "feature_",
        seq_len(nrow(sig))
      )
    ),
    all(
      nonempty(sig$feature_name)
    ),
    all(
      is_uniprot_accession(
        sig$feature_name
      )
    ),
    is.numeric(sig$score),
    all(
      is.finite(sig$score)
    ),
    is.factor(sig$group_label),
    level_check,
    all(
      !is.na(sig$group_label)
    ),
    all(
      as.character(
        sig$group_label
      ) %in% expected_levels
    )
  )

  message(
    context,
    ": ",
    nrow(sig),
    " validated signature rows."
  )

  invisible(TRUE)
}

validate_omic_signature <- function(
    object,
    expected_levels,
    context
) {
  stopifnot(
    !is.null(object$signature),
    is.data.frame(object$signature),
    is.null(object$difexp)
  )

  ## Pre-construction signature tables are required to preserve the
  ## explicitly defined factor-level order. The OmicSignature constructor
  ## may reconstruct factors in a different order while retaining the same
  ## biological labels. Post-construction validation therefore requires
  ## exact label-set preservation rather than exact factor-level ordering.
  validate_signature_table(
    object$signature,
    expected_levels,
    context,
    require_exact_level_order = FALSE
  )

  invisible(TRUE)
}
```

# Build 13 age-associated protein signatures

```{r build-dep-signatures}
DEP_LEVELS <- c(
  "Increasing with age",
  "Decreasing with age"
)

make_dep_osig <- function(tissue_name) {
  dep_audit <- s2_dep_tbl |>
    dplyr::filter(
      Tissue == tissue_name
    ) |>
    dplyr::mutate(
      feature_name = clean_uniprot(
        feature_name
      ),
      score = as.numeric(
        Spearman_r
      ),
      group_label = factor(
        dplyr::case_when(
          score > 0 ~ "Increasing with age",
          score < 0 ~ "Decreasing with age",
          TRUE ~ NA_character_
        ),
        levels = DEP_LEVELS
      )
    )

  stopifnot(
    nrow(dep_audit) > 0L,
    all(
      dep_audit$P_value < 0.05
    ),
    all(
      dep_audit$Type[
        dep_audit$score > 0
      ] == "Upregulated"
    ),
    all(
      dep_audit$Type[
        dep_audit$score < 0
      ] == "Downregulated"
    )
  )

  sig_tbl <- dep_audit |>
    dplyr::filter(
      match_status %in%
        c(
          "mapped",
          "one_to_many"
        ),
      nonempty(feature_name),
      is.finite(P_value),
      P_value < 0.05,
      is.finite(score),
      score != 0,
      !is.na(group_label)
    ) |>
    dplyr::arrange(
      P_value,
      dplyr::desc(
        abs(score)
      ),
      source_row,
      source_accession_index
    ) |>
    dplyr::transmute(
      feature_name,
      score,
      group_label
    ) |>
    as.data.frame()

  sig_tbl <- sigrepo_standardize_signature(
    sig_tbl,
    context = paste(
      tissue_name,
      "age-associated protein signature"
    )
  )

  validate_signature_table(
    sig_tbl,
    DEP_LEVELS,
    paste(
      tissue_name,
      "age-associated protein signature"
    )
  )

  stopifnot(
    all(
      as.character(
        sig_tbl$group_label[
          sig_tbl$score > 0
        ]
      ) == "Increasing with age"
    ),
    all(
      as.character(
        sig_tbl$group_label[
          sig_tbl$score < 0
        ]
      ) == "Decreasing with age"
    )
  )

  meta <- OmicSignature::createMetadata(
    signature_name = paste0(
      "Aging_Hs_",
      name_token_map[[tissue_name]],
      "_AgeAssociatedProteins_Ding2025"
    ),
    organism = "Homo sapiens",
    direction_type = "bi-directional",
    assay_type = ASSAY_STR,
    phenotype = paste0(
      "Age-associated proteins in ",
      tissue_name
    ),
    platform = PLATFORM_STR,
    sample_type = sample_type_map[[tissue_name]],
    covariates = COVARIATES_STR,
    p_value_cutoff = 0.05,
    keywords = sigrepo_keywords(
      c(
        "aging",
        tissue_name,
        "proteomics",
        "age-associated proteins"
      )
    ),
    author = AUTHOR,
    PMID = PMID_STR,
    year = YEAR_NUM,
    description = paste0(
      "Age-associated proteomic signature for human ",
      tissue_name,
      " from Table S2 of Ding et al. Cell 2025. Positive Spearman ",
      "scores indicate increasing abundance with age and negative scores ",
      "indicate decreasing abundance. Membership uses the exact reported ",
      "criterion P_value < 0.05. Multi-accession source cells remain ",
      "expanded. The worksheet is an already selected result table, so ",
      "difexp is NULL."
    )
  )

  OmicSignature::OmicSignature$new(
    metadata = meta,
    signature = sig_tbl,
    difexp = NULL
  )
}

dep_oSig_list <- purrr::set_names(
  dep_tissues
) |>
  purrr::map(
    make_dep_osig
  )

OmSC_DEP <- OmicSignature::OmicSignatureCollection$new(
  OmicSigList = dep_oSig_list,
  metadata = list(
    collection_name =
      "Aging_Hs_MultiTissue_AgeAssociatedProteins_Ding2025",
    description = paste0(
      "Tissue-specific age-associated proteomic signatures across ",
      "13 human tissues from Table S2 of Ding et al. Cell 2025."
    ),
    author = AUTHOR
  )
)
```

# Build 6 proteome-transcriptome decoupling signatures

```{r build-decoupling-signatures}
DECOUPLING_LEVELS <-
  "Age-related proteome-transcriptome decoupling"

make_decoupling_osig <- function(tissue_name) {
  source_rows <- s2_dec_mapped |>
    dplyr::filter(
      Tissue == tissue_name
    )

  stopifnot(
    nrow(source_rows) > 0L,
    all(
      source_rows$R_Young > 0
    ),
    all(
      source_rows$Diff < 0
    ),
    all(
      abs(
        source_rows$Diff -
          (
            source_rows$R_Old -
              source_rows$R_Young
          )
      ) < 1e-6
    )
  )

  sig_tbl <- source_rows |>
    dplyr::filter(
      match_status %in%
        c(
          "mapped",
          "one_to_many"
        ),
      nonempty(feature_name)
    ) |>
    dplyr::mutate(
      score = abs(
        as.numeric(Diff)
      ),
      group_label = factor(
        DECOUPLING_LEVELS,
        levels = DECOUPLING_LEVELS
      )
    ) |>
    dplyr::filter(
      is.finite(score),
      score > 0
    ) |>
    dplyr::arrange(
      dplyr::desc(score),
      source_row,
      feature_name
    ) |>
    dplyr::transmute(
      feature_name,
      score,
      group_label
    ) |>
    as.data.frame()

  sig_tbl <- sigrepo_standardize_signature(
    sig_tbl,
    context = paste(
      tissue_name,
      "proteome-transcriptome decoupling signature"
    )
  )

  validate_signature_table(
    sig_tbl,
    DECOUPLING_LEVELS,
    paste(
      tissue_name,
      "proteome-transcriptome decoupling signature"
    )
  )

  meta <- OmicSignature::createMetadata(
    signature_name = paste0(
      "Aging_Hs_",
      name_token_map[[tissue_name]],
      "_ProteomeTranscriptomeDecoupling_Ding2025"
    ),
    organism = "Homo sapiens",
    direction_type = "uni-directional",
    assay_type = ASSAY_STR,
    phenotype = paste0(
      "Age-related proteome-transcriptome decoupling in ",
      tissue_name
    ),
    platform = PLATFORM_STR,
    sample_type = sample_type_map[[tissue_name]],
    covariates = COVARIATES_STR,
    keywords = sigrepo_keywords(
      c(
        "aging",
        tissue_name,
        "proteomics",
        "proteome-transcriptome decoupling"
      )
    ),
    author = AUTHOR,
    PMID = PMID_STR,
    year = YEAR_NUM,
    description = paste0(
      "Proteome-transcriptome decoupling signature for human ",
      tissue_name,
      " from Table S2 of Ding et al. Cell 2025. The study defines ",
      "decoupling proteins as proteins with positive protein-mRNA ",
      "correlation in young samples and reduced correlation during aging. ",
      "The reported change is R_Old - R_Young and is negative for these ",
      "rows; score is abs(Diff), representing the magnitude of coupling ",
      "loss. Valid one-to-many UniProt mappings remain expanded. The ",
      "worksheet is an already selected result table, so difexp is NULL."
    )
  )

  OmicSignature::OmicSignature$new(
    metadata = meta,
    signature = sig_tbl,
    difexp = NULL
  )
}

dec_oSig_list <- purrr::set_names(
  dec_tissues
) |>
  purrr::map(
    make_decoupling_osig
  )

OmSC_Decoupling <-
  OmicSignature::OmicSignatureCollection$new(
    OmicSigList = dec_oSig_list,
    metadata = list(
      collection_name =
        "Aging_Hs_MultiTissue_ProteomeTranscriptomeDecoupling_Ding2025",
      description = paste0(
        "Tissue-specific proteome-transcriptome decoupling ",
        "signatures across 6 human tissues from Table S2 of ",
        "Ding et al. Cell 2025."
      ),
      author = AUTHOR
    )
  )
```

# Build 13 proteomic aging-clock feature signatures

```{r build-clock-signatures}
CLOCK_LEVELS <-
  "Proteomic aging clock feature"

make_clock_osig <- function(tissue_name) {
  source_rows <- s3_clock_mapped |>
    dplyr::filter(
      Tissue == tissue_name
    )

  stopifnot(
    nrow(source_rows) > 0L,
    all(
      is.finite(source_rows$Importance)
    )
  )

  sig_tbl <- source_rows |>
    dplyr::filter(
      match_status %in%
        c(
          "mapped",
          "one_to_many"
        ),
      nonempty(feature_name),
      Importance >
        CLOCK_IMPORTANCE_CUTOFF
    ) |>
    dplyr::mutate(
      score = as.numeric(
        Importance
      ),
      group_label = factor(
        CLOCK_LEVELS,
        levels = CLOCK_LEVELS
      )
    ) |>
    dplyr::filter(
      is.finite(score),
      score >
        CLOCK_IMPORTANCE_CUTOFF
    ) |>
    dplyr::arrange(
      dplyr::desc(score),
      source_row,
      feature_name
    ) |>
    dplyr::transmute(
      feature_name,
      score,
      group_label
    ) |>
    as.data.frame()

  sig_tbl <- sigrepo_standardize_signature(
    sig_tbl,
    context = paste(
      tissue_name,
      "proteomic aging-clock signature"
    )
  )

  validate_signature_table(
    sig_tbl,
    CLOCK_LEVELS,
    paste(
      tissue_name,
      "proteomic aging-clock signature"
    )
  )

  stopifnot(
    all(
      sig_tbl$score >
        CLOCK_IMPORTANCE_CUTOFF
    )
  )

  meta <- OmicSignature::createMetadata(
    signature_name = paste0(
      "Aging_Hs_",
      name_token_map[[tissue_name]],
      "_ProteomicClockFeatures_Ding2025"
    ),
    organism = "Homo sapiens",
    direction_type = "uni-directional",
    assay_type = ASSAY_STR,
    phenotype = paste0(
      "Proteomic aging clock features for ",
      tissue_name
    ),
    platform = PLATFORM_STR,
    sample_type = sample_type_map[[tissue_name]],
    covariates = COVARIATES_STR,
    score_cutoff =
      CLOCK_IMPORTANCE_CUTOFF,
    keywords = sigrepo_keywords(
      c(
        "aging",
        tissue_name,
        "proteomics",
        "proteomic aging clock"
      )
    ),
    author = AUTHOR,
    PMID = PMID_STR,
    year = YEAR_NUM,
    description = paste0(
      "Proteomic aging-clock feature signature for human ",
      tissue_name,
      " from Table S3 of Ding et al. Cell 2025. Features were ",
      "identified using permutation importance. Membership applies ",
      "the exact Figure S5A criterion Importance > 1, and the ",
      "reported Importance value is retained as score. Valid ",
      "one-to-many UniProt mappings remain expanded. The worksheet ",
      "is a selected feature-importance table, so difexp is NULL."
    )
  )

  OmicSignature::OmicSignature$new(
    metadata = meta,
    signature = sig_tbl,
    difexp = NULL
  )
}

clock_oSig_list <- purrr::set_names(
  clock_tissues
) |>
  purrr::map(
    make_clock_osig
  )

OmSC_Clock <-
  OmicSignature::OmicSignatureCollection$new(
    OmicSigList = clock_oSig_list,
    metadata = list(
      collection_name =
        "Aging_Hs_MultiTissue_ProteomicClockFeatures_Ding2025",
      description = paste0(
        "Tissue-specific proteomic aging-clock feature signatures ",
        "across 13 human tissues from Table S3 of Ding et al. ",
        "Cell 2025, using permutation importance > 1."
      ),
      author = AUTHOR
    )
  )
```

# Final object and collection validation

```{r final-validation}
stopifnot(
  length(dep_oSig_list) == 13L,
  length(dec_oSig_list) == 6L,
  length(clock_oSig_list) == 13L
)

purrr::iwalk(
  dep_oSig_list,
  function(obj, nm) {
    validate_omic_signature(
      obj,
      DEP_LEVELS,
      paste(
        nm,
        "DEP OmicSignature"
      )
    )
  }
)

purrr::iwalk(
  dec_oSig_list,
  function(obj, nm) {
    validate_omic_signature(
      obj,
      DECOUPLING_LEVELS,
      paste(
        nm,
        "decoupling OmicSignature"
      )
    )
  }
)

purrr::iwalk(
  clock_oSig_list,
  function(obj, nm) {
    validate_omic_signature(
      obj,
      CLOCK_LEVELS,
      paste(
        nm,
        "clock OmicSignature"
      )
    )
  }
)

dep_members <- sigrepo_collection_members(
  OmSC_DEP
)

dec_members <- sigrepo_collection_members(
  OmSC_Decoupling
)

clock_members <- sigrepo_collection_members(
  OmSC_Clock
)

stopifnot(
  length(dep_members) == 13L,
  length(dec_members) == 6L,
  length(clock_members) == 13L
)

extract_signature_names <- function(x) {
  vapply(
    x,
    function(object) {
      as.character(
        object$metadata$signature_name
      )
    },
    character(1)
  )
}

stopifnot(
  setequal(
    extract_signature_names(
      dep_members
    ),
    vapply(
      dep_oSig_list,
      function(object) {
        as.character(
          object$metadata$signature_name
        )
      },
      character(1)
    )
  ),
  setequal(
    extract_signature_names(
      dec_members
    ),
    vapply(
      dec_oSig_list,
      function(object) {
        as.character(
          object$metadata$signature_name
        )
      },
      character(1)
    )
  ),
  setequal(
    extract_signature_names(
      clock_members
    ),
    vapply(
      clock_oSig_list,
      function(object) {
        as.character(
          object$metadata$signature_name
        )
      },
      character(1)
    )
  )
)

ding_signature_counts <- dplyr::bind_rows(
  tibble::tibble(
    family = "Age-associated proteins",
    tissue = names(dep_oSig_list),
    signature_rows = purrr::map_int(
      dep_oSig_list,
      ~ nrow(.x$signature)
    )
  ),
  tibble::tibble(
    family = "Proteome-transcriptome decoupling",
    tissue = names(dec_oSig_list),
    signature_rows = purrr::map_int(
      dec_oSig_list,
      ~ nrow(.x$signature)
    )
  ),
  tibble::tibble(
    family = "Proteomic aging clock",
    tissue = names(clock_oSig_list),
    signature_rows = purrr::map_int(
      clock_oSig_list,
      ~ nrow(.x$signature)
    )
  )
)

print(ding_signature_counts)

EXPECTED_DEP_SIGNATURE_ROWS <-
  EXPECTED_DEP_SOURCE_ROWS

EXPECTED_DECOUPLING_SIGNATURE_ROWS <-
  EXPECTED_DECOUPLING_SOURCE_ROWS

EXPECTED_CLOCK_SIGNATURE_ROWS <-
  EXPECTED_CLOCK_SOURCE_ROWS

actual_signature_rows <- function(
    count_table,
    family_name,
    expected_tissues
) {
  current <- count_table |>
    dplyr::filter(
      family == family_name
    )

  stats::setNames(
    current$signature_rows,
    current$tissue
  )[
    expected_tissues
  ]
}

stopifnot(
  identical(
    actual_signature_rows(
      ding_signature_counts,
      "Age-associated proteins",
      names(
        EXPECTED_DEP_SIGNATURE_ROWS
      )
    ),
    EXPECTED_DEP_SIGNATURE_ROWS
  ),
  identical(
    actual_signature_rows(
      ding_signature_counts,
      "Proteome-transcriptome decoupling",
      names(
        EXPECTED_DECOUPLING_SIGNATURE_ROWS
      )
    ),
    EXPECTED_DECOUPLING_SIGNATURE_ROWS
  ),
  identical(
    actual_signature_rows(
      ding_signature_counts,
      "Proteomic aging clock",
      names(
        EXPECTED_CLOCK_SIGNATURE_ROWS
      )
    ),
    EXPECTED_CLOCK_SIGNATURE_ROWS
  )
)

message(
  paste0(
    "Ding 2025 development validation passed for ",
    nrow(ding_signature_counts),
    " OmicSignature objects in three collections, with all expected ",
    "per-signature row counts confirmed."
  )
)
```

# Approval status

No unresolved items remain for this approved reference workflow.

The workflow uses the reviewed and frozen Ding 2025 source-label-to-UniProt
mapping cache. It passed source-table validation, controlled-vocabulary
validation, mapping validation, expected-count validation, UniProt-format
validation, probe-ID validation, pre-construction and post-construction
signature validation, collection validation, and a clean end-to-end render.

The approved workflow creates 13 age-associated protein signatures,
6 proteome-transcriptome decoupling signatures, and 13 proteomic aging-clock
signatures, organized into three OmicSignature collections.

# Optional save step

```{r save, eval=FALSE}
save_osig_list <- function(
    osig_list,
    outdir
) {
  purrr::iwalk(
    osig_list,
    function(object, object_key) {
      output_file <- file.path(
        outdir,
        paste0(
          object$metadata$signature_name,
          "_oSig.rds"
        )
      )

      if (file.exists(output_file)) {
        stop(
          "Refusing to overwrite existing file: ",
          output_file,
          call. = FALSE
        )
      }

      saveRDS(
        object,
        output_file
      )
    }
  )
}

save_osig_list(
  dep_oSig_list,
  OUTDIR
)

save_osig_list(
  dec_oSig_list,
  OUTDIR
)

save_osig_list(
  clock_oSig_list,
  OUTDIR
)

collection_outputs <- c(
  DEP = file.path(
    OUTDIR,
    "Aging_Hs_MultiTissue_AgeAssociatedProteins_Ding2025_OmSC.rds"
  ),
  Decoupling = file.path(
    OUTDIR,
    "Aging_Hs_MultiTissue_ProteomeTranscriptomeDecoupling_Ding2025_OmSC.rds"
  ),
  Clock = file.path(
    OUTDIR,
    "Aging_Hs_MultiTissue_ProteomicClockFeatures_Ding2025_OmSC.rds"
  )
)

stopifnot(
  !any(
    file.exists(
      collection_outputs
    )
  )
)

saveRDS(
  OmSC_DEP,
  collection_outputs[["DEP"]]
)

saveRDS(
  OmSC_Decoupling,
  collection_outputs[["Decoupling"]]
)

saveRDS(
  OmSC_Clock,
  collection_outputs[["Clock"]]
)
```
