---
title: "Derivation of EMT signatures - Khalil Kass Youssef et. al"
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
    toc_float: true
---

```{r global, echo=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

```{r}
current_input <- knitr::current_input(
  dir = TRUE
)

if (
  length(current_input) == 1L &&
    !is.na(current_input) &&
    nzchar(current_input)
) {
  RMD_DIR <- dirname(
    normalizePath(
      current_input,
      winslash = "/",
      mustWork = TRUE
    )
  )
} else {
  RMD_DIR <- normalizePath(
    getwd(),
    winslash = "/",
    mustWork = TRUE
  )
}

REFERENCE_DIR <- file.path(
  RMD_DIR,
  "reference_data"
)

OUTDIR <- file.path(
  RMD_DIR,
  "outputs"
)

INFILE <- Sys.getenv(
  "YOUSSEF2024_INPUT_FILE",
  unset = ""
)

if (!nzchar(INFILE)) {
  stop(
    paste0(
      "YOUSSEF2024_INPUT_FILE is not set. ",
      "Set it to the Youssef 2024 supplementary workbook."
    ),
    call. = FALSE
  )
}

stopifnot(
  file.exists(INFILE),
  dir.exists(REFERENCE_DIR)
)

```

```{r}
library(Biobase)
library(dplyr)
library(readr)
library(readxl)
library(tibble)
library(tidyr)

# ------------------------------------------------------------------------------
# user_lib <- "/restricted/projectnb/montilab-p/personal/hnikouei/R/x86_64-pc-linux-gnu-library/4.4"
# dir.create(user_lib, recursive = TRUE, showWarnings = FALSE)

# .libPaths(c(user_lib, .libPaths()))
# 
# if (!requireNamespace("remotes", quietly = TRUE)) {
#   install.packages("remotes", lib = user_lib, repos = "https://cloud.r-project.org")
# }
# 
# remotes::install_github(
#   "montilab/OmicSignature",
#   lib = user_lib,
#   upgrade = "never",
#   dependencies = TRUE
# )

# .libPaths(c(user_lib, .libPaths()))
# library(OmicSignature)
# ------------------------------------------------------------------------------

user_lib <- Sys.getenv(
  "SIGREPO_R_LIB",
  unset = ""
)

if (nzchar(user_lib)) {
  stopifnot(
    dir.exists(user_lib)
  )

  .libPaths(
    c(
      user_lib,
      .libPaths()
    )
  )
}

library(OmicSignature)


find_codex_reference_root <- function(start = getwd()) {
  current <- normalizePath(
    start,
    winslash = "/",
    mustWork = TRUE
  )

  repeat {
    candidate <- file.path(current, "codex_reference")

    if (dir.exists(candidate)) {
      return(
        normalizePath(
          candidate,
          winslash = "/",
          mustWork = TRUE
        )
      )
    }

    parent <- dirname(current)

    if (identical(parent, current)) {
      stop(
        "Could not locate the codex_reference directory.",
        call. = FALSE
      )
    }

    current <- parent
  }
}

CODEX_REFERENCE_ROOT <- find_codex_reference_root(
  RMD_DIR
)

OMIC_HELPER <- file.path(
  CODEX_REFERENCE_ROOT,
  "helpers",
  "omicsignature_compat.R"
)

stopifnot(file.exists(OMIC_HELPER))
source(
  OMIC_HELPER,
  local = TRUE
)

OMIC_API <- sigrepo_assert_compatible_api()
print(OMIC_API)
```

# Preflight interpretation

## Study and scope

- Study: Youssef et al., Nature Cancer, 2024.
- PMID: 39414946.
- Requested scope: the human epithelial and mesenchymal EMT-component
  gene sets listed in Supplementary Table 1.
- Supplementary Table 1 states that these component lists were extracted
  from prior references and used by the current paper to score breast
  cancer cell lines.

## Biologically justified objects

1. Epithelial EMT-component membership signature.
2. Mesenchymal EMT-component membership signature.
3. One collection containing exactly those two related objects.

The BC-PINGs, validation, pathway, and other supplementary analyses are
not separate objects in this workflow.

## Source classification and object design

- Source: Supplementary Table 1.
- Source type: curated membership-only gene lists.
- Score: constant membership weight of 1.
- Direction type: uni-directional for each object.
- `difexp`: `NULL`; the source is not a broader differential-analysis table.
- Source identifiers: human gene symbols.
- Repository feature identifiers: Ensembl gene IDs from Ensembl release 114.
- Mapping: HGNC symbol to Ensembl v114, retaining valid one-to-many mappings.
- Unmatched rows remain in the mapping audit and are excluded only from
  object input.
- No genuine assay probe identifier is supplied. Probe IDs are assigned once
  through the tested OmicSignature compatibility helper.

## Resolved metadata

- Approved `sample_type`: `breast cancer cell line`.
- Approved platform: `transcriptomics by bulk RNA-seq`.
- Both values are validated against the controlled-vocabulary terms exposed
  by the compatible installed OmicSignature package.


# Load data (read Supplementary Table 1)
and extract raw human gene symbols
```{r}
## 1) Read Supplementary Table 1
## Supplementary Table 1 contains:
##   - epithelial genes in columns 1:7
##   - mesenchymal genes in columns 8:11

tbl1 <- readxl::read_excel(
  path = INFILE,
  sheet = "Supplementary Table 1",
  skip = 3,
  col_names = FALSE
)

names(tbl1) <- paste0("col", seq_len(ncol(tbl1)))

## Preserve the source row and source column for every supplied gene symbol.
## The +3 accounts for the rows skipped above.

epi_raw <- tbl1 |>
  dplyr::mutate(source_row = dplyr::row_number() + 3L) |>
  dplyr::select(source_row, col1:col7) |>
  tidyr::pivot_longer(
    cols = col1:col7,
    names_to = "source_column",
    values_to = "original_feature_name"
  ) |>
  dplyr::mutate(
    source_file = basename(INFILE),
    source_sheet = "Supplementary Table 1",
    original_feature_name = trimws(as.character(original_feature_name)),
    signature_class = "Epithelial"
  ) |>
  dplyr::filter(
    !is.na(original_feature_name),
    nzchar(original_feature_name)
  ) |>
  dplyr::distinct(
    source_file,
    source_sheet,
    source_row,
    source_column,
    original_feature_name,
    signature_class
  )

mes_raw <- tbl1 |>
  dplyr::mutate(source_row = dplyr::row_number() + 3L) |>
  dplyr::select(source_row, col8:col11) |>
  tidyr::pivot_longer(
    cols = col8:col11,
    names_to = "source_column",
    values_to = "original_feature_name"
  ) |>
  dplyr::mutate(
    source_file = basename(INFILE),
    source_sheet = "Supplementary Table 1",
    original_feature_name = trimws(as.character(original_feature_name)),
    signature_class = "Mesenchymal"
  ) |>
  dplyr::filter(
    !is.na(original_feature_name),
    nzchar(original_feature_name)
  ) |>
  dplyr::distinct(
    source_file,
    source_sheet,
    source_row,
    source_column,
    original_feature_name,
    signature_class
  )

cat("Number of epithelial source entries:", nrow(epi_raw), "\n")
cat("Number of mesenchymal source entries:", nrow(mes_raw), "\n")

head(epi_raw)
head(mes_raw)

```

# Map symbols to Ensembl v114 and build the mapping audit
```{r mapping}
## 4) Load the reviewed Symbol -> Ensembl v114 mapping cache
## Template used:
##   Symbol -> Ensembl v114 with reviewed 1:N expansion

MAPPING_CACHE <- file.path(
  REFERENCE_DIR,
  "Youssef2024_HGNC_to_Ensembl_v114.rds"
)

if (!file.exists(MAPPING_CACHE)) {
  stop(
    paste0(
      "The reviewed Ensembl release 114 ",
      "mapping cache is missing: ",
      MAPPING_CACHE
    ),
    call. = FALSE
  )
}

symbol_map <- readRDS(
  MAPPING_CACHE
)

stopifnot(
  is.data.frame(symbol_map),

  identical(
    names(symbol_map),
    c(
      "original_feature_name",
      "feature_name"
    )
  ),

  nrow(symbol_map) == 398L,

  dplyr::n_distinct(
    symbol_map$original_feature_name
  ) == 396L,

  !anyNA(symbol_map),

  all(
    nzchar(
      symbol_map$original_feature_name
    )
  ),

  all(
    stringr::str_detect(
      symbol_map$feature_name,
      "^ENSG[0-9]+$"
    )
  ),

  !anyDuplicated(symbol_map)
)

message(
  paste0(
    "Loaded the reviewed Youssef 2024 ",
    "HGNC-to-Ensembl release 114 mapping cache: ",
    nrow(symbol_map),
    " mapping rows for ",
    dplyr::n_distinct(
      symbol_map$original_feature_name
    ),
    " source symbols."
  )
)

## -----------------------------------------------------------------------------
## 5) Apply mapping and retain a complete mapping audit

make_symbol_mapping_audit <- function(source_tbl, symbol_map) {
  source_tbl |>
    dplyr::left_join(
  symbol_map,
  by = "original_feature_name",
  relationship = "many-to-many"
) |>
    dplyr::group_by(
      source_file,
      source_sheet,
      source_row,
      source_column,
      original_feature_name,
      signature_class
    ) |>
    dplyr::mutate(
      primary_mapping_key = original_feature_name,
      rescue_mapping_key = NA_character_,
      mapping_source = "reviewed static HGNC-to-Ensembl release 114 mapping cache",
      mapping_multiplicity = dplyr::n_distinct(
        feature_name[!is.na(feature_name) & nzchar(feature_name)]
      ),
      match_status = dplyr::case_when(
        mapping_multiplicity == 0L ~ "unmatched",
        mapping_multiplicity == 1L ~ "mapped",
        mapping_multiplicity > 1L ~ "one_to_many"
      )
    ) |>
    dplyr::ungroup()
}

epi_mapped <- make_symbol_mapping_audit(epi_raw, symbol_map)
mes_mapped <- make_symbol_mapping_audit(mes_raw, symbol_map)

## Optional draft audit writes. Keep FALSE in the reusable reference example.
do_write_audit <- FALSE

if (do_write_audit) {
  saveRDS(
    epi_mapped,
    file = file.path(OUTDIR, "EMT_Epithelial_symbol_to_ensembl_v114_mapped.rds")
  )

  saveRDS(
    mes_mapped,
    file = file.path(OUTDIR, "EMT_Mesenchymal_symbol_to_ensembl_v114_mapped.rds")
  )

  write.csv(
    epi_mapped,
    file = file.path(OUTDIR, "EMT_Epithelial_symbol_to_ensembl_v114_mapped.csv"),
    row.names = FALSE
  )

  write.csv(
    mes_mapped,
    file = file.path(OUTDIR, "EMT_Mesenchymal_symbol_to_ensembl_v114_mapped.csv"),
    row.names = FALSE
  )
}

dir.create(
  OUTDIR,
  recursive = TRUE,
  showWarnings = FALSE
)
```

```{r mapping-qc}
cat("Epithelial source entries:", nrow(epi_raw), "\n")
print(table(epi_mapped$match_status, useNA = "ifany"))

cat("\nMesenchymal source entries:", nrow(mes_raw), "\n")
print(table(mes_mapped$match_status, useNA = "ifany"))

epi_unmatched <- epi_mapped |>
  dplyr::filter(match_status == "unmatched")

mes_unmatched <- mes_mapped |>
  dplyr::filter(match_status == "unmatched")

epi_one_to_many <- epi_mapped |>
  dplyr::filter(match_status == "one_to_many")

mes_one_to_many <- mes_mapped |>
  dplyr::filter(match_status == "one_to_many")

print(epi_unmatched)
print(mes_unmatched)
print(epi_one_to_many)
print(mes_one_to_many)
```
The runtime mapping tables above are authoritative. An increase from source
rows to mapped rows is expected when a source symbol has multiple valid
Ensembl v114 mappings. Valid one-to-many mappings remain expanded. Unmatched
and one-to-many rows must be reviewed in the audit before approval.

The workflow continues with the in-memory `epi_mapped` and `mes_mapped`
objects; it does not reload legacy mapped RDS files.

## Reviewed mapping decisions

The frozen mapping cache was constructed using Ensembl release 114 gene
records and the HGNC 2025-04-01 archive.

Direct Ensembl release 114 symbol mappings were preferred. Supported HGNC
previous-symbol mappings were used as rescue mappings. The ambiguous alias
`CD`, which could refer to either `CTLA4` or `NOD2`, was excluded. For `MALL`
and `SLC6A10P`, the Ensembl IDs supported by HGNC were retained. Nineteen
source symbols remained unresolved after the supported rescue procedure and
were excluded from object construction while remaining documented in the
mapping review records.

# Build the OmicSignature objects 
build signature tables from the mapped output
```{r}
## 7) Build final membership-only signature tables
##
## These are curated membership signatures:
##   - difexp = NULL
##   - score = 1
##   - direction_type = "uni-directional"
##
## Unmatched rows remain in the mapping audit but do not enter the objects.
## Valid one-to-many mappings remain expanded.

make_membership_signature <- function(mapping_audit, label) {
  raw_signature <- mapping_audit |>
    dplyr::filter(
      match_status %in% c("mapped", "one_to_many"),
      !is.na(feature_name),
      nzchar(feature_name)
    ) |>
    dplyr::transmute(
      feature_name = trimws(feature_name),
      score = 1,
      group_label = factor(label, levels = label)
    ) |>
    as.data.frame()

  ## No genuine assay/probe identifier is supplied by the source.
  ## Standardize once and stop if the package removes a legitimate mapping row.
  sigrepo_standardize_signature(
    raw_signature,
    context = paste(label, "EMT membership signature")
  )
}
# The helper verifies that row membership, repeated feature mappings, andautomatically assigned probe IDs remain valid.


sig_epithelial <- make_membership_signature(
  mapping_audit = epi_mapped,
  label = "Epithelial"
)

sig_mesenchymal <- make_membership_signature(
  mapping_audit = mes_mapped,
  label = "Mesenchymal"
)

validate_membership_signature <- function(sig, expected_label) {
  stopifnot(
    all(c("probe_id", "feature_name", "score", "group_label") %in% names(sig)),
    is.character(sig$probe_id),
    all(!is.na(sig$probe_id) & nzchar(sig$probe_id)),
    !anyDuplicated(sig$probe_id),
    all(!is.na(sig$feature_name) & nzchar(sig$feature_name)),
    is.numeric(sig$score),
    all(is.finite(sig$score)),
    all(sig$score == 1),
    is.factor(sig$group_label),
    identical(levels(sig$group_label), expected_label),
    all(as.character(sig$group_label) == expected_label)
  )
}

validate_membership_signature(sig_epithelial, "Epithelial")
validate_membership_signature(sig_mesenchymal, "Mesenchymal")

cat("Final epithelial signature rows:", nrow(sig_epithelial), "\n")
cat("Final mesenchymal signature rows:", nrow(sig_mesenchymal), "\n")

EXPECTED_EPITHELIAL_ROWS <- 272L
EXPECTED_MESENCHYMAL_ROWS <- 126L

stopifnot(
  nrow(sig_epithelial) ==
    EXPECTED_EPITHELIAL_ROWS,

  nrow(sig_mesenchymal) ==
    EXPECTED_MESENCHYMAL_ROWS,

  identical(
    sig_epithelial$probe_id,
    paste0(
      "feature_",
      seq_len(EXPECTED_EPITHELIAL_ROWS)
    )
  ),

  identical(
    sig_mesenchymal$probe_id,
    paste0(
      "feature_",
      seq_len(EXPECTED_MESENCHYMAL_ROWS)
    )
  )
)

message(
  paste0(
    "Expected EMT signature counts confirmed: ",
    EXPECTED_EPITHELIAL_ROWS,
    " epithelial and ",
    EXPECTED_MESENCHYMAL_ROWS,
    " mesenchymal rows."
  )
)
## -----------------------------------------------------------------------------
## 9) Metadata
# These are curated EMT membership lists, not DE results.

## These two values must be supported by the source references and must exactly
## match the terms returned by the compatible installed OmicSignature package.
##
## Do not approve this reference while either value remains unresolved.
SAMPLE_STR <- "breast cancer cell line"

PLATFORM_STR <-
  "transcriptomics by bulk RNA-seq"


sigrepo_assert_sample_type(
  SAMPLE_STR
)

sigrepo_assert_platform(
  PLATFORM_STR
)


meta_epithelial <- OmicSignature::createMetadata(
  signature_name = "BreastCancer_Hs_BreastCancerCellLines_EpithelialEMTComponent_Youssef2024", # Epithelial-to-mesenchymal transition.
  organism       = "Homo sapiens",
  direction_type = "uni-directional",
  assay_type     = "transcriptomics",
  phenotype      = "Epithelial EMT component",
  platform       = PLATFORM_STR,
  sample_type    = SAMPLE_STR,
  covariates     = "none",
  keywords       = sigrepo_keywords(
    c("EMT", "epithelial", "breast cancer", "cell line")
  ),
  author         = "Khalil Kass Youssef",
  PMID           = "39414946",
  year           = 2024,
  description    = "Epithelial EMT-component gene set from Supplementary Table 1, used in Youssef et al. Nature Cancer 2024 to score the epithelial side of the EMT spectrum across human breast cancer cell lines."
)

meta_mesenchymal <- OmicSignature::createMetadata(
  signature_name = "BreastCancer_Hs_BreastCancerCellLines_MesenchymalEMTComponent_Youssef2024",
  organism       = "Homo sapiens",
  direction_type = "uni-directional",
  assay_type     = "transcriptomics",
  phenotype      = "Mesenchymal EMT component",
 platform       = PLATFORM_STR,
  sample_type    = SAMPLE_STR,
  covariates     = "none",
  keywords       = sigrepo_keywords(
    c("EMT", "mesenchymal", "breast cancer", "cell line")
  ),
  author         = "Khalil Kass Youssef",
  PMID           = "39414946",
  year           = 2024,
  description    = "Mesenchymal EMT-component gene set from Supplementary Table 1, used in Youssef et al. Nature Cancer 2024 to score the mesenchymal side of the EMT spectrum across human breast cancer cell lines."
)

```

```{r}
## 10) Create OmicSignature objects
## difexp = NULL because these are list-only signatures

oSig_epithelial <- OmicSignature::OmicSignature$new(
  metadata  = meta_epithelial,
  signature = sig_epithelial,
  difexp    = NULL
)

oSig_mesenchymal <- OmicSignature::OmicSignature$new(
  metadata  = meta_mesenchymal,
  signature = sig_mesenchymal,
  difexp    = NULL
)

```

post-construction validation
```{r}
validate_created_omic_signature <-
  function(object, object_name) {

    stopifnot(
      !is.null(object$signature),
      is.data.frame(object$signature),
      "probe_id" %in%
        names(object$signature),
      is.character(
        object$signature$probe_id
      ),
      !anyNA(
        object$signature$probe_id
      ),
      all(
        nzchar(
          object$signature$probe_id
        )
      ),
      !anyDuplicated(
        object$signature$probe_id
      )
    )

    message(
      object_name,
      ": ",
      nrow(object$signature),
      " valid signature rows."
    )

    invisible(TRUE)
  }

validate_created_omic_signature(
  oSig_epithelial,
  "Youssef epithelial EMT"
)

validate_created_omic_signature(
  oSig_mesenchymal,
  "Youssef mesenchymal EMT"
)
```

# Collection
```{r}
## 11) Create collection

OmSC_EMT <- OmicSignature::OmicSignatureCollection$new(
  OmicSigList = list(
    epithelial  = oSig_epithelial,
    mesenchymal = oSig_mesenchymal
  ),
  metadata = list(
    collection_name = "BreastCancer_Hs_BreastCancerCellLines_EMTComponents_Youssef2024",
    description     = "Human epithelial and mesenchymal EMT-component gene sets from Supplementary Table 1 of Youssef et al. Nature Cancer 2024, used to score EMT state across human breast cancer cell lines.",
    author          = "Khalil Kass Youssef"
  )
)

# collection validation
emt_members <- sigrepo_collection_members(OmSC_EMT)

stopifnot(
  length(emt_members) == 2L
)

actual_signature_names <- vapply(
  emt_members,
  function(x) {
    as.character(
      x$metadata$signature_name
    )
  },
  character(1)
)

expected_signature_names <- c(
  as.character(
    oSig_epithelial$metadata$signature_name
  ),
  as.character(
    oSig_mesenchymal$metadata$signature_name
  )
)

stopifnot(
  setequal(
    actual_signature_names,
    expected_signature_names
  )
)

message(
  paste0(
    "EMT collection contains the ",
    "expected two OmicSignature objects."
  )
)
```

# Approval status

No unresolved items remain for this approved reference workflow.

The workflow uses a reviewed and frozen Ensembl release 114 mapping cache.
It passed controlled-vocabulary validation, mapping validation, expected-count
validation, probe-ID validation, pre-construction and post-construction
validation, collection validation, and a clean end-to-end render.

# Save
```{r save, eval=FALSE}
## 12) Optional save step
## Keep eval=FALSE in the reusable reference example.

# Do not execute the save chunk while this file is being used as a reference example.

saveRDS(
  oSig_epithelial,
  file = file.path(
    OUTDIR,
    "EMT_Hs_BreastCancerCellLines_EpithelialEMTComponent_Youssef2024_oSig.rds"
  )
)

saveRDS(
  oSig_mesenchymal,
  file = file.path(
    OUTDIR,
    "EMT_Hs_BreastCancerCellLines_MesenchymalEMTComponent_Youssef2024_oSig.rds"
  )
)

saveRDS(
  OmSC_EMT,
  file = file.path(
    OUTDIR,
    "EMT_Hs_BreastCancerCellLines_EpithelialMesenchymalComponents_Youssef2024_OmSC.rds"
  )
)


## 13) Optional quick inspection

head(oSig_epithelial$signature)
head(oSig_mesenchymal$signature)

nrow(oSig_epithelial$signature)
nrow(oSig_mesenchymal$signature)
```
