---
title: "Annotate signatures with hypeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Annotate signatures with hypeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The code runs only when a SigRepo account is supplied through the
# SIGREPO_TEST_DB_USER and SIGREPO_TEST_DB_PASSWORD environment variables;
# otherwise the tutorial shows its code without running it. Credentials never
# go in this file. Everything here only reads from the database.
has_account <- base::nzchar(base::Sys.getenv("SIGREPO_TEST_DB_USER")) &&
  base::nzchar(base::Sys.getenv("SIGREPO_TEST_DB_PASSWORD"))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.align = "center",
  eval = has_account
)
```

This tutorial will walk you through how to fetch, and annotate signatures with gene set enrichment, in three
steps:

1. Find the signatures you want to test with `searchSignature()`.
2. Test them against a geneset collection with `runHypeR()`.
3. Plot the results with `plotHypeRDots()` and `plotHypeREnrichment()`.

The same analysis runs without code in the SigRepo website's Annotate tab; see
[the end of this tutorial](#the-same-analysis-in-the-annotate-tab).

# Load packages:
All of the required packages for annotate (fgsea, hypeR) are built into SigRepo.

```{r load packages}
library(SigRepo)
```

# Connect to SigRepo

Connect with your own SigRepo account.

```{r connect, eval = FALSE}
conn_handler <- SigRepo::newConnHandler(
  # required: your SigRepo account
  user = "montilab",
  password = "sigrepo",
  dbname = "sigrepo",
  host = "sigrepo.org",
  port = 3306
)
```

```{r account, include = FALSE}
conn_handler <- SigRepo::newConnHandler(
  dbname = base::Sys.getenv("SIGREPO_TEST_DB_NAME", "sigrepo"),
  host = base::Sys.getenv("SIGREPO_TEST_DB_HOST", "sigrepo.org"),
  port = base::as.numeric(base::Sys.getenv("SIGREPO_TEST_DB_PORT", "3306")),
  user = base::Sys.getenv("SIGREPO_TEST_DB_USER"),
  password = base::Sys.getenv("SIGREPO_TEST_DB_PASSWORD"),
  api_host = base::Sys.getenv("SIGREPO_TEST_API_HOST", "http://142.93.67.157:8020"),
  api_port = base::as.numeric(base::Sys.getenv("SIGREPO_TEST_API_PORT", "8020"))
)
```

# Step 1: Find the signatures to test

`searchSignature()` returns one row per signature you can see, with its
metadata. Here we look for human signatures:

```{r search}
human <- SigRepo::searchSignature(
  # required:
  conn_handler = conn_handler,

  # optional filters, any combination:
  organism = "Homo sapiens",
  phenotype = NULL,
  sample_type = NULL,
  platform_name = NULL,
  signature_name = NULL,
  signature_id = NULL,
  user_name = NULL,
  verbose = FALSE
)

nrow(human)
```

Narrow the results down to the sample you want. Enrichment works best on
signatures stored with their full differential expression table (`has_difexp`),
because the ranked tests (KS and fgsea) rank every measured gene. Smaller difexp
tables that already carry gene symbols run fastest. Here we take the same
experiment in two breast cancer cell lines: knockdown of the aryl hydrocarbon
receptor (AhR) in MDA-MB-231 and in SUM149PT cells, each compared with control
cells (`CN`):

```{r sample}
experiments <- c(
  "BreastCancer_Hs_MDA_AhR_KD_vs_CN_MontiLab",
  "BreastCancer_Hs_SUM149PT_AhR_KD_vs_CN_MontiLab"
)

my_signatures <- human[
  human$signature_name %in% experiments & human$has_difexp == 1,
  c("signature_id", "signature_name", "assay_type", "num_of_difexp", "num_up_regulated", "num_down_regulated")
]
my_signatures
```

# Step 2: Run the enrichment

For enrichment, SigRepo uses [hypeR](https://github.com/montilab/hypeR), an
open-source R package for geneset enrichment developed by the Monti lab. You can
read its documentation at
[montilab.github.io/hypeR-docs](https://montilab.github.io/hypeR-docs/).
`runHypeR()` turns SigRepo signatures into hypeR's input, runs the test in your R
session, and returns hypeR's own result objects.

`runHypeR()` tests every signature against a geneset collection and returns a
hypeR result. Four things are required:

- **which signatures**: `signature_id` (or `signature_name`) with `conn_handler`. Also takes OmicSignature objects read from `.rds` files, as `omic_signature`;
- **`organism`**: the organism of every signature in the call;
- **`test`**: `"hypergeometric"`, `"kstest"` or `"fgsea"`;
- **`genesets`**: `"msigdb"` with an MSigDB collection, or your own genesets.

## The three tests

| `test` | Asks | Tests |
|---|---|---|
| `"hypergeometric"` | Are the signature's genes over-represented in a geneset? | the signature's genes, one query per `group_label` |
| `"kstest"` | Do a geneset's genes sit toward the top or the bottom of the ranking? | every measured gene, ranked by score, then by the negated score |
| `"fgsea"` | Do a geneset's genes sit toward the top or the bottom of the ranking? | every measured gene, ranked by score, with permutation p-values |

The hypergeometric test only needs the signature's genes. The ranked tests
(KS and fgsea) need signatures stored with their difexp table, rank all of its
genes in one list rather than splitting them by `group_label`, and always test
both ends of that list: each ranked result has a `| up` query (genesets toward
the highest scores) and a `| down` query (toward the lowest). The sections
below run each test in turn.

## Hypergeometric test

```{r hypergeometric}
hyp <- SigRepo::runHypeR(
  # required:
  conn_handler = conn_handler,
  signature_id = my_signatures$signature_id,  # or signature_name = my_signatures$signature_name
  organism = "Homo sapiens",
  test = "hypergeometric",
  genesets = "msigdb",
  msigdb_collection = "H",                    # required with genesets = "msigdb": "H", "C2", "C5", ...

  # optional:
  msigdb_subcollection = NULL,                # e.g. "CP:REACTOME" with msigdb_collection = "C2"
  split = TRUE,                               # one query per group_label of each signature (here AhR / CN)
  background = NULL,                          # NULL: the signature's measured genes; or a number or gene vector
  min_query_genes = 4,                        # skip a query with fewer genes found in the genesets
  fdr_scope = "run",                          # adjust p-values across the whole call ("query": per query)
  pval = 1,                                   # keep genesets with p-value at or below this
  fdr = 1,                                    # keep genesets with FDR at or below this
  verbose = FALSE
)

names(hyp$data)
head(hyp$data[[1]]$data[, c("label", "pval", "fdr", "overlap", "geneset")])
```

The result is a hypeR `multihyp`: one `hyp` per query in `hyp$data`, each with a
results table in `$data` and the settings it ran with in `$info`. With
`split = TRUE`, each signature gives one query per `group_label`, here `AhR`
and `CN`.

## GSEA with fgsea

`fgsea` tests both ends of the ranking in one run and files each geneset under
`| up` or `| down` by the sign of its enrichment score:


```{r fgsea}
gsea <- SigRepo::runHypeR(
  # required:
  conn_handler = conn_handler,
  signature_id = my_signatures$signature_id,
  organism = "Homo sapiens",
  test = "fgsea",
  genesets = "msigdb",
  msigdb_collection = "H",

  # optional for fgsea:
  power = 1,                                  # weight of each gene's score (fgsea's gseaParam)
  seed = 1,                                   # fgsea samples at random; the seed makes a run repeatable
  fgsea_args = list(minSize = 15, maxSize = 500),  # other fgsea::fgseaMultilevel() settings
  fdr = 1,
  verbose = FALSE
)


# tables are split into up and down tables, in this instance there should be four seperate tables, two for each signature.
names(gsea$data)
# showing the result of one table
head(gsea$data[["BreastCancer_Hs_SUM149PT_AhR_KD_vs_CN_MontiLab | down"]]$data[, c("label", "fdr", "nes", "overlap")], 3)
```

## KS test

hypeR's KS test only finds genesets at the top of a ranking, so `runHypeR()`
runs it twice: on the scores (`| up`) and on the negated scores (`| down`),
adjusting the p-values of both runs for multiple testing together:

```{r kstest}
ks <- SigRepo::runHypeR(
  # required:
  conn_handler = conn_handler,
  signature_id = my_signatures$signature_id[1],
  organism = "Homo sapiens",
  test = "kstest",
  genesets = "msigdb",
  msigdb_collection = "H",

  # optional for kstest:
  score_col = "score",                        # the difexp column that ranks the genes
  power = 1,                                  # 0 for an unweighted ranked test
  verbose = FALSE
)

head(ks$data)
```

## One organism per call

Genesets are made of one organism's gene symbols, so every signature in a call
must match `organism`. Asking for another organism, or mixing human and mouse
signatures, stops with an error that names the signatures. Run each organism
in its own call:

```{r organism mismatch, error = TRUE}
SigRepo::runHypeR(
  conn_handler = conn_handler,
  signature_id = my_signatures$signature_id[1],
  organism = "Mus musculus",
  test = "hypergeometric",
  genesets = "msigdb",
  msigdb_collection = "MH",
  verbose = FALSE
)
```

# Step 3: Visualize the results

Every plot is drawn from the result object and titled with the test that was
run. `plotHypeRDots()` and `plotHypeREnrichment()` return ggplot objects, which
you can save with `ggplot2::ggsave()`.

## Dot plot: every signature at once

One column per query and one row per geneset; a blank cell means the geneset
did not pass the cutoffs for that query. With several signatures, each
signature's columns sit under a short code (`S1`, `S2`, ...) explained in the
caption.

```{r dots hypergeometric, fig.width = 8, fig.height = 6}

# hypergeometric dot plot example

SigRepo::plotHypeRDots(
  # required:
  hyp_obj = hyp,

  # optional:
  top = 15,                                   # number of genesets, the best by FDR across all queries
  fdr = 0.25,                                 # hide genesets above this FDR
  val = "fdr",                                # rank and colour by "fdr" or "pval"
  color_by = "significance",                  # "score" (NES) for kstest and fgsea results
  size_by = "geneset",                        # dot size: "geneset", "overlap" or "none"
  signature_key = TRUE,                       # FALSE labels columns with the full query names
  title = NULL                                # NULL: the test that was run
)

```

For the ranked tests, `color_by = "score"` shows the direction of enrichment:

```{r dots fgsea, fig.width = 8, fig.height = 6}
SigRepo::plotHypeRDots(gsea, top = 15, fdr = 0.05, color_by = "score")
SigRepo::plotHypeRDots(ks, fdr = 0.05, top = 15,  color_by = "score")
```



```{r enrichment hypergeometric, fig.width = 6, fig.height = 4.5}
SigRepo::plotHypeREnrichment(
  hyp_obj = hyp,
  geneset = "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION",
  query = "BreastCancer_Hs_SUM149PT_AhR_KD_vs_CN_MontiLab | AhR"
)
```

# The same analysis in the Annotate tab

The SigRepo website's Annotate tab is a front end to `runHypeR()` and follows the
same rules:

| In the tab | In R |
|---|---|
| Step 1: pick up to 10 signatures, upload OmicSignature `.rds` files, or type gene lists | `signature_id`, `omic_signature` or `signature` |
| Step 2: **Organism (required)**, then MSigDB or custom genesets | `organism`, `genesets` |
| Step 3: **Test (required)**, plus split and minimum genes for the hypergeometric test | `test`, `split`, `min_query_genes` |
| Step 3: Background | `background` |

- A run tests one organism. If a picked or uploaded signature is of another
  organism, the tab names it and will not run until you change the organism or
  remove the signature; previewing the queries still works.
- There is no direction setting. KS and GSEA always rank the whole difexp and
  return an `| up` and a `| down` query for every signature, as in R.
- The R code tab writes the `runHypeR()` call and the plot calls that reproduce
  the run, with `organism` and `test` always included.

See `?runHypeR` and the help page of each plotting function for every argument.
