---
title: "Signature Collection Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Signature Collection Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  fig.path = "./images/", 
  message = TRUE,
  warning = TRUE,
  error = TRUE,
  collapse = TRUE,
  comment = "#>"
)

# Load R packages
# For data cleaning, extraction and manipulation
library(tidyverse)

# For loading and installing packages
library(devtools)

# Load SigRepo package
devtools::load_all()

# Load OmicSignature package
library("OmicSignature")

```

# Installation

- Using `devtools` package

```

# Load devtools package
library(devtools)

# Install SigRepo
devtools::install_github(repo = 'montilab/SigRepo')

# Install OmicSignature
devtools::install_github(repo = 'montilab/OmicSignature')

```

# Load packages 

```
# Load tidyverse package
library(tidyverse)

# Load SigRepo package
library(SigRepo)

# Load OmicSignature package
library(OmicSignature)
```

# Create a handler to connect to SigRepo Database

```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
# Get the test connection handler
conn_handler <- SigRepo:::global_var[["test_conn_handler"]]
```

```
# Create a connection handler
conn_handler <- SigRepo::newConnHandler(
  dbname = "sigrepo", 
  host = "sigrepo.org", 
  port = 3306, 
  user = <your_username>, 
  password = <your_password>
)
```

# Load signatures

Here, we provided 4 signature objects that came with the package for demonstrations:

1. omic_signature_1 (`Myc_reduce_mice_liver_24m_v1`)
2. omic_signature_2 (`Myc_reduce_mice_liver_24m_v2`)
3. omic_signature_3 (`Myc_reduce_mice_liver_24m_v3`)
4. omic_signature_4 (`Myc_reduce_mice_liver_24m_v4`)

```{r}
# Read in the signature objects
utils::data("omic_signature_1", package = "SigRepo")
utils::data("omic_signature_2", package = "SigRepo")
utils::data("omic_signature_3", package = "SigRepo")
utils::data("omic_signature_4", package = "SigRepo")
```

# Create an omic collection 

Here, we will create a collection with two signatures, `omic_signature_1` and `omic_signature_1`, provided above. See `?OmicSignatureCollection()` from `OmicSignature` package for more details on how to create an omic collection object.

```{r}
# Create a metadata object for the collection
metadata <- base::list(
  "collection_name" = "my_collection",
  "description" = "An example of signature collection"
)

# Create an omic collection using OmicSignatureCollection() from OmicSignature package
omic_collection <- OmicSignature::OmicSignatureCollection$new(
  OmicSigList = base::list(omic_signature_1, omic_signature_2),
  metadata = metadata
)
```

# Upload a collection to the database

The `SigRepo::addCollection()` function allows users to upload a collection to the database. 

__IMPORTANT NOTE:__ 

- The user `MUST HAVE` an `editor` or `admin` account to use this function. 
- A collection `MUST BE` an R6 object obtained from `OmicSignature::OmicSignatureCollection()`

```{r}
SigRepo::addCollection(
  conn_handler = conn_handler,        # A handler contains user credentials to establish connection to a remote database
  omic_collection = omic_collection,  # An R6 object obtained from OmicSignature::OmicSignatureCollection()
  visibility = FALSE,                 # Whether to make the collection public or private. Default is FALSE.
  return_collection_id = FALSE,       # Whether to return the uploaded collection id. Default is FALSE.
  verbose = TRUE                      # Whether to print diagnostic messages. Default is TRUE.
)
```

# Search for a collection in the database

The `SigRepo::searchCollection()` function allows users to retrieve all collections that are available in the database. 

## Example 1: Search for all collections 

```{r}
# Get all collections
collection_tbl_1 <- SigRepo::searchCollection(conn_handler = conn_handler)

# WARNINGS: THE COLLECTION TABLE CAN BE LARGE, SO ONLY THE FIRST SIX OBSERVATIONS ARE SHOWN.
if(base::nrow(collection_tbl_1) > 0){
  knitr::kable(
    utils::head(collection_tbl_1), 
    row.names = FALSE
  )
}
```

## Example 2: Search for a specific collection by its name, e.g., `collection_name = "my_collection"`.

```{r}
collection_tbl_2 <- SigRepo::searchCollection(
  conn_handler = conn_handler, 
  collection_name = "my_collection"
)

if(base::nrow(collection_tbl_2) > 0){
  knitr::kable(
    collection_tbl_2, 
    row.names = FALSE
  )
}
```

# Retrieve a set of collections in the database

The `SigRepo::getCollection()` function allows users to retrieve a set of collections that are `PUBLICLY` available in the database. 

__IMPORTANT NOTE:__ 

- Users can `ONLY RETRIEVE` a list of collections that are publicly available in the database including their own uploaded collections. 
- If a collection is `PRIVATE` and belongs to other user in the database, users will need to be given an `editor` permission from its owner to access, retrieve, and edit their collections.

## Example 1: Get all collections that are publicly available in the database 

```
# WARNINGS: THE COLLECTION LIST CAN BE LARGE TO DOWNLOAD (NOT RUN)
collection_list_1 <- SigRepo::getCollection(conn_handler = conn_handler)
```

## Example 2: Get a specific collection that is publicly available or owned by user in the database, e.g., `collection_name = "my_collection"`

```{r}
collection_list_2 <- SigRepo::getCollection(
  conn_handler = conn_handler,
  collection_name = "my_collection"
)
```

# Update a collection metadata

The `SigRepo::updateCollectionMetadata()` function allows users to update the metadata information of a collection in the database. 

__IMPORTANT NOTE:__ 

- Users `MUST HAVE` an `editor` or `admin` account to use this function. 
- Furthermore, users can `ONLY UPDATE` their own uploaded signatures or were given an `editor` permission from its owner to access, retrieve, and edit their collections.

For example, if you want to change the description of `"my_collection"` in the database to `"This is the updated description."` You can use the `SigRepo::updateCollectionMetadata()` function as follows:

```{r}
# Let's search for my_collection in the database
collection_tbl <- SigRepo::searchCollection(
  conn_handler = conn_handler, 
  collection_name = "my_collection"
)

# If the collection exists, update it with the new description
if(base::nrow(collection_tbl) > 0){
  SigRepo::updateCollectionMetadata(
    conn_handler = conn_handler, 
    collection_id = collection_tbl$collection_id,
    description =  "This is the updated description." 
  )
}
```

Now, let's search for `"my_collection"` in the database and see if its description has been updated.

```{r}
# Search for my_collection in the database
collection_tbl <- SigRepo::searchCollection(
  conn_handler = conn_handler, 
  collection_name = "my_collection"
)

# If the collection exists, display the table
if(base::nrow(collection_tbl) > 0){
  knitr::kable(
    collection_tbl,
    row.names = FALSE
  )
}
```

# Add signatures to an existing collection in the database

The `SigRepo::addSignatureToCollection()` function allows users to add a set of signatures to an existing collection in the database. 

__IMPORTANT NOTE:__ 

- Users `MUST HAVE `an `editor` or `admin` account to use this function. 
- Furthermore, users can `ONLY UPDATE` their own uploaded signatures or were given an `editor` permission from its owner to access, retrieve, and edit their collections.

For example, if you wish to create two new signatures (`"omic_signature_3"`, `"omic_signature_4"`) and add them to `"my_collection" `in the database.

```{r}
# Add omic_signature_3 to the database
signature_id_3 <- SigRepo::addSignature(
  conn_handler = conn_handler, 
  omic_signature = omic_signature_3,
  return_signature_id = TRUE,
  verbose = FALSE
)

# Add omic_signature_4 to the database
signature_id_4 <- SigRepo::addSignature(
  conn_handler = conn_handler, 
  omic_signature = omic_signature_4,
  return_signature_id = TRUE,
  verbose = FALSE
)

# Search for my_collection in the database
collection_tbl <- SigRepo::searchCollection(
  conn_handler = conn_handler, 
  collection_name = "my_collection",
  verbose = FALSE
)

# If both collection and signatures exist, add its signature ids to collection.
if(base::nrow(collection_tbl) > 0 && base::length(signature_id_3) > 0 && base::length(signature_id_4) > 0){
  SigRepo::addSignatureToCollection(
    conn_handler = conn_handler, 
    collection_id = collection_tbl$collection_id,
    signature_id =  c(signature_id_3, signature_id_4)
  )
}
```

Now, let's search for `"my_collection"` in the database and see if the signatures are added to the collection.

```{r}
# Search for my_collection in the database
collection_tbl <- SigRepo::searchCollection(
  conn_handler = conn_handler, 
  collection_name = "my_collection"
)

# If collection exists, display the table
if(base::nrow(collection_tbl) > 0){
  knitr::kable(
    collection_tbl,
    row.names = FALSE
  )
}
```

# Remove a list of signatures from a collection in the database

The `SigRepo::removeSignatureFromCollection()` function allows users to remove a list of signatures from a collection in the database. 

__IMPORTANT NOTE:__ 

- Users `MUST HAVE `an `editor` or `admin` access to use this function. 
- Furthermore, users can `ONLY UPDATE `their own uploaded signatures or were given an `editor` permission from its owner to access, retrieve, and edit their collections.

For example, if you wish to remove `"Myc_reduce_mice_liver_24m_v1"` from `"my_collection" `in the database

```{r}
# Lets' check if signature = Myc_reduce_mice_liver_24m_v1 is in my_collection
collection_tbl <- SigRepo::searchCollection(
  conn_handler = conn_handler, 
  collection_name = "my_collection",
  signature_name = "Myc_reduce_mice_liver_24m_v1"
)

# If signature collection combo exists, remove the signature.
if(base::nrow(collection_tbl) > 0){
  SigRepo::removeSignatureFromCollection(
    conn_handler = conn_handler, 
    collection_id = collection_tbl$collection_id,
    signature_id = collection_tbl$signature_id
  )
}
```

Now, let's search for `"my_collection"` in the database and see if signatures have been removed from the collection.

```{r}
# Search for my_collection in the database
collection_tbl <- SigRepo::searchCollection(
  conn_handler = conn_handler, 
  collection_name = "my_collection"
)

# If collection exists, display the table
if(base::nrow(collection_tbl) > 0){
  knitr::kable(
    collection_tbl,
    row.names = FALSE
  )
}
```

# Delete the whole collection from the database

The `SigRepo::deleteCollection()` function allows users to delete a collection from the database. 

__IMPORTANT NOTE:__ 

- Users `MUST HAVE` an `editor` or `admin` account to use this function. 
- Furthermore, users can `ONLY DELETE` their own uploaded collections or were given an `editor` permission from other users in the database to access and delete their collection.
- A collection can `ONLY BE REMOVED OR DELETED` one at a time.

```{r}
# Search for my_collection in the database 
collection_tbl <- SigRepo::searchCollection(
  conn_handler = conn_handler, 
  collection_name = "my_collection"
)

# If the collection exists, remove it from the database
if(base::nrow(collection_tbl) > 0){
  SigRepo::deleteCollection(
    conn_handler = conn_handler, 
    collection_id = collection_tbl$collection_id
  )
}
```

Finally, let's search for `"my_collection"` in the database and see if the collection has been removed.

```{r}
collection_tbl <- SigRepo::searchCollection(
  conn_handler = conn_handler, 
  collection_name = "my_collection"
)
```