---
title: "Compare signatures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The code runs only when a SigRepo account is supplied through the
# SIGREPO_TEST_DB_USER and SIGREPO_TEST_DB_PASSWORD environment variables;
# otherwise the tutorial shows its code without running it. Credentials never
# go in this file. Everything here only reads from the database.
has_account <- base::nzchar(base::Sys.getenv("SIGREPO_TEST_DB_USER")) &&
  base::nzchar(base::Sys.getenv("SIGREPO_TEST_DB_PASSWORD"))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.align = "center",
  eval = has_account
)
```

This tutorial will walk you through how to fetch signatures and compare them
with each other, in three steps:

1. Find the signatures you want to compare with `searchSignature()`.
2. Compare them with `compareSignatures()`.
3. Plot the results with `OmicSignature::signature_similarity_heatmap()`.

The same analysis runs without code in the SigRepo website's Compare tab; see
[the end of this tutorial](#the-same-analysis-in-the-compare-tab).

# Load packages:
`compareSignatures()` runs the comparison with
[OmicSignature](https://github.com/montilab/OmicSignature), which is installed
with SigRepo. Two parts of this tutorial need extra packages: the GSEA method
needs fgsea, and the heatmaps need ComplexHeatmap and circlize.

```{r install, eval = FALSE}
# only if they are not installed yet:
BiocManager::install(c("fgsea", "ComplexHeatmap"))
install.packages("circlize")
```

```{r load packages}
library(SigRepo)
```

# Connect to SigRepo

Connect with your own SigRepo account.

```{r connect, eval = FALSE}
conn_handler <- SigRepo::newConnHandler(
  # required: your SigRepo account
  user = "montilab",
  password = "sigrepo",
  dbname = "sigrepo",
  host = "sigrepo.org",
  port = 3306
)
```

```{r account, include = FALSE}
conn_handler <- SigRepo::newConnHandler(
  dbname = base::Sys.getenv("SIGREPO_TEST_DB_NAME", "sigrepo"),
  host = base::Sys.getenv("SIGREPO_TEST_DB_HOST", "sigrepo.org"),
  port = base::as.numeric(base::Sys.getenv("SIGREPO_TEST_DB_PORT", "3306")),
  user = base::Sys.getenv("SIGREPO_TEST_DB_USER"),
  password = base::Sys.getenv("SIGREPO_TEST_DB_PASSWORD"),
  api_host = base::Sys.getenv("SIGREPO_TEST_API_HOST", "http://142.93.67.157:8020"),
  api_port = base::as.numeric(base::Sys.getenv("SIGREPO_TEST_API_PORT", "8020"))
)
```

# Step 1: Find the signatures to compare

`searchSignature()` returns one row per signature you can see, with its
metadata. Here we look for human signatures:

```{r search}
human <- SigRepo::searchSignature(
  # required:
  conn_handler = conn_handler,

  # optional filters, any combination:
  organism = "Homo sapiens",
  phenotype = NULL,
  sample_type = NULL,
  platform = NULL,
  signature_name = NULL,
  signature_id = NULL,
  user_name = NULL,
  verbose = FALSE
)

nrow(human)
```

Narrow the results down to the signatures you want to compare. Every method can
compare signatures by their genes, but the ranked methods (KS and GSEA) also
need a bi-directional signature stored with its full differential expression
table (`has_difexp`) to rank against. Here we take two knockdowns in two breast
cancer cell lines, MDA-MB-231 and SUM149PT: knockdown of the aryl hydrocarbon
receptor (AhR), and knockdown of CYP1B1, a gene AhR switches on. Each is
compared with control cells (`CN`):

```{r sample}
experiments <- c(
  "BreastCancer_Hs_MDA_AhR_KD_vs_CN_MontiLab",
  "BreastCancer_Hs_SUM149PT_AhR_KD_vs_CN_MontiLab",
  "BreastCancer_Hs_MDA_CYP1B1_KD_vs_CN_MontiLab",
  "BreastCancer_Hs_SUM149PT_CYP1B1_KD_vs_CN_MontiLab"
)

my_signatures <- human[
  human$signature_name %in% experiments & human$has_difexp == 1,
  c("signature_id", "signature_name", "type", "num_of_difexp", "num_up_regulated", "num_down_regulated")
]
my_signatures
```

# Step 2: Run the comparison

`compareSignatures()` fetches the signatures from SigRepo and compares them with
`OmicSignature::compare_omic_signatures()`, in your R session. Only the
signatures are required:

- **which signatures**: `signature_ids` (or `signature_names`) with `conn_handler`. Also takes OmicSignature objects, as `omic_signatures`;
- **a second list** (optional): `signature_ids2`, `signature_names2` or `omic_signatures2`;
- **`method`** (optional): `"overlap"` (the default), `"ks_rank"`, `"ks_score"` or `"gsea"`.

## One list or two

| Signatures given | Compares | Result matrices |
|---|---|---|
| a first list only | every signature with every other one; needs at least two | square: the first list as rows and as columns |
| a first and a second list | each signature in the first list with each one in the second; one per list is enough | the first list as rows, the second list as columns |

## The four methods

| `method` | Asks | Tests |
|---|---|---|
| `"overlap"` | Do two signatures share more genes than chance? | Fisher's exact test on the two gene sets, with their Jaccard similarity |
| `"ks_rank"` | Do one signature's genes sit toward one end of the other's ranking? | where the genes fall in the ranking (KS test) |
| `"ks_score"` | Do one signature's genes sit toward one end of the other's ranking? | the genes' ranking scores against every other gene's (two-sample KS test) |
| `"gsea"` | Do one signature's genes sit toward one end of the other's ranking? | fgsea's enrichment score, with permutation p-values |

Every method starts by choosing each signature's genes, one set per
`group_label`: the genes with `adj_p` at or below `adj_p_cutoff` and an
absolute `score` at or above `score_cutoff`, best `adj_p` first, topped up to
`min_features` and cut at `max_feature`.

The ranked methods also rank the second list (or, with one list, the other
signature of each pair) by its difexp table: every measured gene, ordered by
`-log10(p_value)`, with one `group_label`'s genes at the top and the other's at
the bottom. That is why the ranking side must be bi-directional with a difexp
table; the first list can hold any signature. The sections below run each
method in turn.

## Overlap

```{r overlap, results = "hide"}
overlap <- SigRepo::compareSignatures(
  # required: at least two signatures when there is no second list
  conn_handler = conn_handler,
  signature_ids = my_signatures$signature_id,   # or signature_names = my_signatures$signature_name

  # optional:
  method = "overlap",
  adj_p_cutoff = 0.05,                          # keep genes with adj_p at or below this
  score_cutoff = 0,                             # and an absolute score at or above this
  min_features = 5,                             # top a set up to this many genes (at least 3)
  max_feature = 500,                            # keep at most this many genes per group_label
  background = NULL,                            # NULL: every gene in the compared signatures
  alternative = "greater",                      # test for more shared genes than chance
  adjust = FALSE,                               # TRUE: adjust the p-values within each matrix
  p_adjust_method = "BH",
  verbose = FALSE
)
```

```{r overlap structure}
names(overlap)
names(overlap$comparisons)
overlap$label_order
```

The result is a list: the `method`, the `comparisons`, the `label_order` and
the `background` it used. A bi-directional signature is compared one
`group_label` at a time: `level1_vs_level1` compares every signature's first
label with the others' first label, and `level2_vs_level2` the second labels.
`label_order` shows which labels those are. Here `level1` is `CN` for every
signature: the genes higher in control cells, which go down after the
knockdown.

For the overlap method each comparison holds three matrices: `jaccard`,
`pvalue`, and `counts`, the number of shared genes (on the diagonal, each
signature's own number of genes). Rows and columns are named by the
signatures' names:

```{r overlap result}
names(overlap$comparisons$level1_vs_level1)
overlap$comparisons$level1_vs_level1$counts[
  "BreastCancer_Hs_SUM149PT_AhR_KD_vs_CN_MontiLab",
  "BreastCancer_Hs_SUM149PT_CYP1B1_KD_vs_CN_MontiLab"
]
```

## Fetch once, compare many times

Every call with `signature_ids` or `signature_names` fetches the signatures
again. To run several comparisons, fetch them once with `getSignature()` and
pass them as `omic_signatures`. Each signature takes its name in the results
from its list name, so this is also the place to give them short names:

```{r fetch, results = "hide"}
signatures <- SigRepo::getSignature(
  conn_handler = conn_handler,
  signature_id = my_signatures$signature_id,
  verbose = FALSE
)
```

```{r short names}
short_names <- c(
  BreastCancer_Hs_MDA_AhR_KD_vs_CN_MontiLab = "MDA_AhR",
  BreastCancer_Hs_SUM149PT_AhR_KD_vs_CN_MontiLab = "SUM149PT_AhR",
  BreastCancer_Hs_MDA_CYP1B1_KD_vs_CN_MontiLab = "MDA_CYP1B1",
  BreastCancer_Hs_SUM149PT_CYP1B1_KD_vs_CN_MontiLab = "SUM149PT_CYP1B1"
)
names(signatures) <- short_names[names(signatures)]
names(signatures)
```

OmicSignature objects saved as `.rds` files work the same way:
`omic_signatures = list(my_signature = readRDS("my_signature.rds"))`. Objects
and database signatures can be mixed in one list.

## Choose the labels with label_pairing

Levels are paired by position, using each signature's own `group_label` order.
To choose them, give both labels for each signature, the first label becoming
`level1`. Here we put the knockdown first, so that `level1_vs_level1` compares
the genes that go up after each knockdown:

```{r label pairing}
pairing <- list(
  MDA_AhR = c("AhR", "CN"),
  SUM149PT_AhR = c("AhR", "CN"),
  MDA_CYP1B1 = c("CYP1B1", "CN"),
  SUM149PT_CYP1B1 = c("CYP1B1", "CN")
)

overlap <- SigRepo::compareSignatures(
  # required:
  omic_signatures = signatures,                 # no conn_handler needed for objects

  # optional:
  method = "overlap",
  label_pairing = pairing,                      # names are the signature names in the result
  verbose = FALSE
)

overlap$label_order
round(overlap$comparisons$level1_vs_level1$jaccard, 3)
```

When signatures left to position disagree on their label order, the comparison
warns that the levels may not line up.

## KS test: rank position

With two lists, each signature in the first list (the rows) is tested against
the ranking of each signature in the second list (the columns). Here we ask
whether each knockdown in MDA-MB-231 looks like the knockdowns in SUM149PT:

```{r ks rank}
ks <- SigRepo::compareSignatures(
  # required:
  omic_signatures = signatures[c("MDA_AhR", "MDA_CYP1B1")],           # the rows: the genes tested
  omic_signatures2 = signatures[c("SUM149PT_AhR", "SUM149PT_CYP1B1")], # the columns: the rankings

  # optional:
  method = "ks_rank",
  label_pairing = pairing[c("MDA_AhR", "MDA_CYP1B1")],
  label_pairing2 = pairing[c("SUM149PT_AhR", "SUM149PT_CYP1B1")],
  p_value_col = "p_value",                      # the difexp column that ranks the genes
  alternative = "greater",                      # test for genes toward the top of the ranking
  verbose = FALSE
)

names(ks$comparisons$level1_vs_level1)
lapply(ks$comparisons, function(x) round(x$score, 3))
lapply(ks$comparisons, function(x) signif(x$pvalue, 2))
```

The ranked methods return a `score` and a `pvalue` matrix per comparison. In
`level1_vs_level1`, each ranking has the column signature's `level1` genes at
the top. A positive score means the row signature's `level1` genes sit toward
that top, where they agree with it; a negative score means they sit toward the
bottom, among the column signature's `level2` genes. The p-value tests the
`alternative`: with `"greater"`, genes toward the top.

Read the two together. The ranking puts the most significant genes of both
labels at its two ends, so a signature's genes can gather at both ends at once.
The score then reports the stronger end, while the p-value still tests the top:
a negative score with a small p-value means genes toward both ends, more of
them at the bottom.

A ranking signature whose difexp table has no `p_value` column is ranked by its
`pvalue` column instead, or by `adj_p` if it has neither; the comparison warns
when it falls back to `adj_p`.

## KS test: score distribution

`"ks_score"` takes the same arguments. It compares the ranking scores of the
row signature's genes with the scores of every other gene, and signs the result
by which of the two is higher on average:

```{r ks score}
ks_score <- SigRepo::compareSignatures(
  omic_signatures = signatures[c("MDA_AhR", "MDA_CYP1B1")],
  omic_signatures2 = signatures[c("SUM149PT_AhR", "SUM149PT_CYP1B1")],
  method = "ks_score",
  label_pairing = pairing[c("MDA_AhR", "MDA_CYP1B1")],
  label_pairing2 = pairing[c("SUM149PT_AhR", "SUM149PT_CYP1B1")],
  verbose = FALSE
)

lapply(ks_score$comparisons, function(x) round(x$score, 3))
```

## GSEA with fgsea

`"gsea"` runs fgsea on each pair and returns its normalized enrichment score
(NES), positive toward the top of the ranking and negative toward the bottom.
fgsea samples at random, so set a seed to make a run repeatable:

```{r gsea, warning = FALSE}
set.seed(1)
gsea <- SigRepo::compareSignatures(
  # required:
  omic_signatures = signatures[c("MDA_AhR", "MDA_CYP1B1")],
  omic_signatures2 = signatures[c("SUM149PT_AhR", "SUM149PT_CYP1B1")],

  # optional for gsea:
  method = "gsea",
  label_pairing = pairing[c("MDA_AhR", "MDA_CYP1B1")],
  label_pairing2 = pairing[c("SUM149PT_AhR", "SUM149PT_CYP1B1")],
  gsea_score = "NES",                           # the fgsea column returned as the score: "NES", "ES", ...
  minSize = 1,                                  # fgsea's smallest gene set
  maxSize = Inf,                                # fgsea's largest gene set
  nproc = 0,                                    # fgsea workers
  verbose = FALSE
)

lapply(gsea$comparisons, function(x) round(x$score, 2))
```

fgsea warns when many genes share a ranking score, as genes with the same
p-value do; the order of those genes is then arbitrary. The warnings are hidden
above.

## When signatures cannot be compared

A signature that does not exist, or that your account cannot see, is left out
with a warning, and the comparison goes ahead with the rest:

```{r missing signature, results = "hide"}
partial <- SigRepo::compareSignatures(
  conn_handler = conn_handler,
  signature_names = c(my_signatures$signature_name[1:2], "Not_a_signature"),
  verbose = FALSE
)
```

The comparison stops when there is nothing to compare, for instance a single
signature without a second list:

```{r one signature, error = TRUE}
SigRepo::compareSignatures(
  omic_signatures = signatures["MDA_AhR"],
  verbose = FALSE
)
```

The ranked methods also stop when no signature on the ranking side is
bi-directional with a difexp table. Categorical signatures are not supported by
any method.

# Step 3: Visualize the results

`OmicSignature::signature_similarity_heatmap()` draws any result of
`compareSignatures()` as a ComplexHeatmap heatmap:

| `measure` | Colours cells by | For |
|---|---|---|
| `"jaccard"` | Jaccard similarity | `"overlap"` |
| `"pvalue"` | `-log10(p-value)` | every method |
| `"score"` | the KS score or the NES | `"ks_rank"`, `"ks_score"`, `"gsea"` |

| `mode` | Draws |
|---|---|
| `"separate"` | one heatmap per level (`lev1_vs_lev1`, `lev2_vs_lev2`) |
| `"combined"` | both levels in each cell: `level1` in the bottom-left triangle, `level2` in the top-right |
| `"split"` | `level1` above the diagonal and `level2` below it; one-list overlap results only |

## Overlap heatmaps

A one-list comparison is symmetric, so only one triangle of the matrix is
drawn. Signatures are clustered, so similar ones sit next to each other:

```{r heatmap separate, fig.width = 8, fig.height = 4}
OmicSignature::signature_similarity_heatmap(
  # required:
  comparison = overlap,

  # optional:
  measure = "jaccard",
  mode = "separate",
  triangle = "upper",                           # the triangle drawn for one-list results
  cluster_method = "ward.D",                    # how rows and columns are clustered
  draw = TRUE                                   # FALSE: return the heatmap without drawing it
)
```

```{r heatmap combined, fig.width = 6, fig.height = 4.5}
OmicSignature::signature_similarity_heatmap(overlap, measure = "pvalue", mode = "combined")
```

## Ranked heatmaps

For the ranked methods, `measure = "score"` shows the direction of each
comparison. Rows are the first list and columns the second:

```{r heatmap ks, fig.width = 6, fig.height = 4}
OmicSignature::signature_similarity_heatmap(ks, measure = "score", mode = "combined")
```

Cells that could not be computed, for a ranking signature without a difexp
table, are grey (`na_style = "grey"`) or hatched (`na_style = "hatch"`).

To save a heatmap, build it with `draw = FALSE` and draw it into a file:

```{r save heatmap, eval = FALSE}
ht <- OmicSignature::signature_similarity_heatmap(overlap, draw = FALSE)

pdf("overlap_heatmap.pdf", width = 8, height = 4)
ComplexHeatmap::draw(ht)
dev.off()
```

# The same analysis in the Compare tab

The SigRepo website's Compare tab is a front end to `compareSignatures()` and
follows the same rules:

| In the tab | In R |
|---|---|
| Step 1: List 1, pick signatures from the repository or upload OmicSignature `.rds` files | `signature_ids`, `omic_signatures` |
| Step 1: **Compare against a second list (List 2)** | `signature_ids2`, `omic_signatures2` |
| Step 2: Method, cutoffs, min and max features, alternative, p-value adjustment | `method`, `score_cutoff`, `adj_p_cutoff`, `min_features`, `max_feature`, `alternative`, `adjust`, `p_adjust_method` |
| Step 2: GSEA settings | `gsea_score`, `minSize`, `maxSize` |
| Step 2: Advanced: background, columns, label pairing | `background`, the `*_col` arguments, `label_pairing`, `label_pairing2` |

- Every setting starts at the `compareSignatures()` default, which its label
  shows.
- Choosing KS or GSEA sets the ranking side's filters to bi-directional
  signatures with a difexp table: List 2, or List 1 when there is no List 2.
- The results show the heatmap (with PNG and PDF downloads), every signature
  pair, the matrices, the label order and the R code that reproduces the run.
  **Download result (.rds)**
  saves the result, which `readRDS()` reads back for
  `OmicSignature::signature_similarity_heatmap()`.

See `?compareSignatures` and `?OmicSignature::signature_similarity_heatmap` for
every argument.
