---
title: "HypeR client review and validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HypeR client review and validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = TRUE,
  message = FALSE
)

library(SigRepo)
library(OmicSignature)

utils::data("LLFS_Aging_Gene_2023", package = "SigRepo", envir = environment())
utils::data("Myc_reduce_mice_liver_24m", package = "SigRepo", envir = environment())
```

## Summary

The hypeR client wrapper in SigRepo was adjusted to better match how the underlying package expects query inputs and how SigRepo signatures are actually stored.

Two concrete problems were identified:

1. The adapter was over-splitting signatures by both `group_label` and direction, which created unnecessarily fragmented hypeR query vectors.
2. Some signatures arrive with feature identifiers (for example `feature_name`) but without a symbol column, while hypeR expects a symbol-based vector for enrichment.

The fix keeps hypeR inputs grouped by `group_label` only and, when needed, maps feature identifiers to `gene_symbol` using the corresponding reference feature table in MySQL. When some rows cannot be mapped, the wrapper warns the user instead of silently dropping the data.

---

## Why the change was needed

### 1. Group-only splitting is the correct contract

The purpose of the hypeR wrapper is to prepare enrichment inputs from SigRepo signature objects, not to duplicate the same signature for every direction subclass.

The previous behavior effectively generated inputs such as:

- `signature_name | group_a_up`
- `signature_name | group_a_dn`
- `signature_name | group_b_up`
- `signature_name | group_b_dn`

This creates a large number of small query vectors and is not aligned with the intended grouping semantics for the wrapper. The current behavior groups only on `group_label`:

- `signature_name | treated`
- `signature_name | control`

This created the expected contract for the client-side workflow while remaining compatible with multi-group signatures.

### 2. Some production signatures do not carry symbols in their `difexp`

A major production issue is that not every SigRepo signature includes a `symbol` or `gene_symbol` column. Some signatures are stored with `feature_name`, `probe_id`, or other feature identifiers instead.

That is a problem for hypeR because the enrichment engine is usually driven by gene symbols. The wrapper therefore resolves symbols via the appropriate reference table:

- `transcriptomics_features`
- `proteomics_features`
- `genetic_variants_features`

If a row cannot be mapped, the wrapper warns the user that some entries were dropped and continues with the successfully mapped subset.

---

## Example signatures used in validation

The package ships with a few signature objects that are useful for testing the wrapper and validating the input normalization logic.

### Example 1: `LLFS_Aging_Gene_2023`

This is a transcriptomics signature included in the package and is a standard example for hypeR preparation.

```{r, eval=TRUE}
utils::data("LLFS_Aging_Gene_2023", package = "SigRepo", envir = environment())

prepared <- SigRepo::prepareHypeRSignatures(
  omic_signature = LLFS_Aging_Gene_2023,
  method = "hypergeo",
  verbose = FALSE
)

names(prepared$signatures)
length(prepared$signatures)
```

This signature is useful because it already behaves like a typical symbol-driven transcriptomics object.

### Example 2: `Myc_reduce_mice_liver_24m`

This signature is useful for exercising the edge case where feature names are present but the data may not yet be symbolized. In the original review, this was one of the signatures used to confirm that the wrapper can tolerate non-symbolized feature identifiers.

```{r, eval=TRUE}
utils::data("Myc_reduce_mice_liver_24m", package = "SigRepo", envir = environment())

prepared <- SigRepo::prepareHypeRSignatures(
  omic_signature = Myc_reduce_mice_liver_24m,
  method = "gsea",
  verbose = FALSE
)

prepared$metadata
```

The key point is that the wrapper should not fail just because `feature_name` is present instead of a ready-to-use symbol vector.

### Example 3: synthetic validation signature with feature mapping fallback

The wrapper itself was also exercised with a synthetic signature that mimics the real production issue: `feature_name` records are present, but the symbol column is missing and must be resolved from the reference table.

```{r, eval=FALSE}
# A minimal example of the actual fallback contract
sig <- OmicSignature::OmicSignature$new(
  metadata = list(
    signature_name = "mapped_symbol_signature",
    assay_type = "transcriptomics",
    phenotype = "test_phenotype",
    organism = "Homo sapiens",
    direction_type = "bi-directional",
    others = list()
  ),
  signature = data.frame(
    feature_name = c("feat_1", "feat_2", "feat_3", "feat_4"),
    feature_id = c(101, 102, 103, 104),
    group_label = factor(c("treated", "treated", "control", "control")),
    stringsAsFactors = FALSE
  ),
  difexp = data.frame(
    feature_name = c("feat_1", "feat_2", "feat_3", "feat_4"),
    score = c(2.2, -1.8, 2.5, -0.9),
    p_value = c(0.01, 0.02, 0.03, 0.04),
    group_label = factor(c("treated", "treated", "control", "control")),
    stringsAsFactors = FALSE
  )
)

prepared <- SigRepo::prepareHypeRSignatures(
  omic_signature = sig,
  method = "gsea",
  split_by_group = TRUE,
  verbose = FALSE
)

prepared$signatures
```

This shape reflects the production pattern that triggered the fix: the difexp table has feature keys, but the wrapper has to resolve them to gene symbols before handing them to hypeR.

---

## Implementation behavior

The final client behavior is designed to be conservative and explicit:

- If `symbol` or `gene_symbol` already exists, use it.
- If the data is feature-based, try to map the feature identifiers to gene symbols from the MySQL reference feature tables.
- If mapping succeeds, continue with the mapped values.
- If mapping fails for some rows, warn and drop only the unmapped entries.
- Do not split by positive/negative direction; split by `group_label` only.

This keeps the wrapper aligned with both the SigRepo object model and the hypeR input contract while preserving user visibility into any data loss due to mapping.

---

## Practical takeaway

The HypeR wrapper is now more robust for real-world SigRepo signatures because it handles both of the common production edge cases:

- multi-group signatures that should not be split by direction
- feature-driven signatures that require symbol resolution before enrichment

This makes the wrapper more reliable, especially when processing live database signatures that do not perfectly match the symbols expected by the hypeR package.
