---
title: "Running hypeR enrichment on SigRepo signatures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running hypeR enrichment on SigRepo signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  eval = requireNamespace("hypeR", quietly = TRUE) && requireNamespace("msigdbr", quietly = TRUE)
)
```

```{r}
library(SigRepo)
library(hypeR)

utils::data("LLFS_Aging_Gene_2023", package = "SigRepo")
sig <- LLFS_Aging_Gene_2023
```

## What `runHypeR()` does

`runHypeR()` turns SigRepo signatures into the query vectors `hypeR::hypeR()`
expects, runs hypeR, and returns hypeR's own object. The arguments from `power`
on are hypeR's, with hypeR's defaults.

### Required and optional arguments

Only two things are required:

- **The signatures to test**, as exactly one of `signature_id` or
  `signature_name` (each with `conn_handler`), `omic_signature` (one object or a
  named list), or `signature` (hypeR-native gene symbols, or a named score
  vector for the ranked tests). These cannot be combined: a run tests one source.
- **`genesets`**: either `"msigdb"` with `msigdb_collection`, or your own
  genesets (a named list, `hypeR::gsets` or `hypeR::rgsets`).

Everything else is optional. Each argument documents its own default, and
leaving it alone gives the conventions below. The arguments that accept `NULL`
read it as "choose for me", not as "nothing": `background = NULL` picks a
background per signature, `query_names = NULL` builds the default query names,
and `seed = NULL` leaves fgsea on the session's random-number state.

The shortest call there is names one signature and one geneset collection:
`runHypeR(omic_signature = sig, genesets = hallmark)`, which the next two
sections build up.

Three defaults follow the enrichment conventions taught in BS831
([hypergeometric](https://montilab.github.io/BS831/articles/docs/HyperEnrichment.html),
[KS](https://montilab.github.io/BS831/articles/docs/KS_Enrichment.html))
rather than plain hypeR, so pass the hypeR values when you need a result that
matches a direct `hypeR::hypeR()` call:

| Argument | Default (BS831) | hypeR's behaviour |
|---|---|---|
| `background` | `NULL`: the signature's measured genes from a complete difexp, else 23467 | `23467` |
| `min_query_genes` | `4` genes found in the genesets (hypergeometric) | `1` |
| `fdr_scope` | `"run"`: FDR across every query and geneset in the call | `"query"` |

The hypeR documentation site (montilab.github.io/hypeR-docs) still shows hypeR
1.7.2. hypeR 2.x renamed some helper arguments, e.g.
`msigdb_gsets(category =, subcategory =)` is now `collection =` /
`subcollection =`. SigRepo needs hypeR 2.0.0 or later, so check the help pages
of your installed hypeR (`?hypeR::msigdb_gsets`) rather than the site.

## Genesets

Fetch an MSigDB collection with `genesets = "msigdb"`:

```{r}
hallmark <- SigRepo::getHypeRGenesets("msigdb", msigdb_collection = "H")
hallmark
```

Or pass your own genesets — a named list, `hypeR::gsets` or `hypeR::rgsets`:

```{r}
my_genesets <- list(
  ribosome_like = c("RPL3", "RPL4", "RPS6"),
  interferon_like = c("ISG15", "IFIT1", "MX1")
)
```

`runHypeR()` accepts either form in `genesets`. The `msigdb_*` arguments only
apply with `genesets = "msigdb"`. `genesets = "msigdb"` queries msigdbr on every
call, so when you run several analyses, fetch the collection once with
`getHypeRGenesets()` and pass the object, as the examples below do. Enrichr
libraries (`hypeR::enrichr_gsets()`) and hypeR's curated hierarchies
(`hypeR::hyperdb_rgsets()`, used by `hyp_hmap()`) pass through the same way.

## Inspect the query vectors

`prepareHypeRSignatures()` shows exactly what will be tested and where the gene
symbols came from:

```{r}
prepared <- SigRepo::prepareHypeRSignatures(omic_signature = sig, verbose = FALSE)
prepared$info
lengths(prepared$signatures)
```

This signature's table has no symbol column, so symbols come from its difexp,
joined on `probe_id`. A bi-directional signature has one `group_label` per arm,
and `split = TRUE` (the default) tests each arm separately. Query names are
always unique — built as `"<label> | <group_label>"` (or just `"<label>"` when
there is no group) and disambiguated as `name (2)`, `name (3)` if two
signatures would otherwise collide. To choose your own names, pass
`query_names`, a function of the `info` table:

```{r}
short <- SigRepo::prepareHypeRSignatures(
  omic_signature = sig,
  query_names = function(info) paste0("LLFS_", info$group_label),
  verbose = FALSE
)
names(short$signatures)
```

## Hypergeometric test

```{r}
hyp <- SigRepo::runHypeR(
  omic_signature = sig,
  genesets = hallmark,
  verbose = FALSE
)
hyp
```

The package's example signature stores only 1,000 difexp genes, so the default
background warns that the difexp looks filtered and uses 23467 instead: a
trimmed difexp is not the set of genes the study measured. As in hypeR, `fdr`
and `pval` default to `1`, so every geneset is kept; pass `fdr = 0.05` to
filter. The FDR is adjusted across both arms together (`fdr_scope = "run"`). The result is a hypeR `multihyp` (one `hyp` per arm),
which SigRepo's plotting functions (see "Plotting results") and hypeR's own
functions accept:

```{r, fig.width = 7, fig.height = 5}
SigRepo::plotHypeRDots(hyp, top = 10)
```

`hyp_show()` takes a single `hyp`, so pass one arm:

```{r, eval = FALSE}
hypeR::hyp_show(hyp$data[[1]])
```

The return type follows the input, as in hypeR. A single signature (one
`OmicSignature`, id or name) gives a `hyp` when it can only build one query:
`split = FALSE`, or a kstest with `direction = "up"` or `"down"` on a signature
that is not categorical. Everything else gives a `multihyp`, even when only one query is left after skips, so
batch code can always index `res$data`.

Each `hyp$info` records where its query came from and how it was run:

```{r}
tail(hyp$data[[1]]$info, 14)
```

## Ranked (KS) test

`test = "kstest"` ranks the full difexp table by `score`, highest first.

- The p-value and FDR come from hypeR's unweighted, one-sided KS test. They do
  not depend on `power` or `absolute`.
- `power` changes only the enrichment `score`: `1` (hypeR's default) weights
  hits by |score|; `0` runs hypeR's ranked signature (gene names in rank
  order), so `Signature Type` is `"ranked"`.
- The test finds genesets enriched toward the top of the ranking only.
  `direction = "down"` ranks by the negated score, so genesets concentrated
  among the most negative scores come first; `direction = "both"` runs both and
  returns `"<label> | up"` and `"<label> | down"`.
- `ks_source = "signature"` ranks the signature table instead of the difexp,
  for signatures stored without one. Only the signature's genes are ranked, so
  the test asks where genesets fall within the signature, not within
  everything measured.
- The ranking must be a signed differential score, such as a t statistic or
  logFC, so that the top of the list is one class and the bottom the other. A
  ranking whose scores are all equal or all one sign is skipped with the reason
  `constant_score` or `unsigned_score`.

hypeR's kstest cannot score a geneset whose hits in a query all score exactly
`0` (with `power != 0`), or one that contains every query gene (at any power).
Such genesets are dropped for that query with a warning, which also leaves them
out of the FDR adjustment. A query with no genesets left is skipped with a
warning.

```{r}
ranked <- SigRepo::runHypeR(
  omic_signature = sig,
  genesets = hallmark,
  test = "kstest",
  direction = "both",
  verbose = FALSE
)
names(ranked$data)
head(ranked$data[["LLFS_Aging_Gene_2023 | down"]]$data)
```

## GSEA with fgsea

hypeR's kstest weights hits only for its enrichment `score`; its p-value is an
unweighted KS test toward the top of the ranking, so `power` never changes
significance. `test = "fgsea"` runs `fgsea::fgseaMultilevel()` on the same
ranking instead: its permutation p-values reflect the score weights (`power` is
fgsea's `gseaParam`), and one run tests both tails. Pathways come back split by
the sign of their enrichment score into `"<query> | up"` and `"<query> | down"`.

```{r, eval = requireNamespace("hypeR", quietly = TRUE) && requireNamespace("msigdbr", quietly = TRUE) && requireNamespace("fgsea", quietly = TRUE)}
gsea <- SigRepo::runHypeR(
  omic_signature = sig,
  genesets = hallmark,
  test = "fgsea",
  fgsea_args = list(minSize = 15, maxSize = 500),
  verbose = FALSE
)
names(gsea$data)
head(gsea$data[["LLFS_Aging_Gene_2023 | up"]]$data[, c("label", "pval", "fdr", "nes", "overlap")])
```

fgsea estimates p-values by random sampling. `seed = 1` (the default) makes a
call reproducible without changing your session's random-number state; a
different `seed` gives slightly different p-values, and `seed = NULL` uses the
session's state. `fgsea_args` passes other `fgseaMultilevel()` options, e.g.
`minSize = 15, maxSize = 500` to test only pathways of typical size. The
results can be coloured by NES:

```{r, fig.width = 7, fig.height = 5, eval = requireNamespace("hypeR", quietly = TRUE) && requireNamespace("msigdbr", quietly = TRUE) && requireNamespace("fgsea", quietly = TRUE)}
SigRepo::plotHypeRDots(gsea, color_by = "score", fdr = 0.25)
```

## Plotting results

SigRepo draws its figures from the result objects, so they work for any
`runHypeR()` result without `plotting = TRUE`:

- `plotHypeRDots()` gives one column per query and one row per geneset. Every
  query shows the same `top` genesets (the best by FDR across all queries), and
  a blank cell means the geneset did not pass the cutoffs for that query.
  `color_by = "score"` shows NES (fgsea) or the enrichment score (kstest).
  Every plot is titled with the test that was run. With several signatures,
  each signature's columns sit together under a short code (`S1`, `S2`, ...)
  labelled by group or direction, and the caption lists the full names
  (`hypeRSignatureKey()` returns the same table); a single signature is named
  in the subtitle. `signature_key = FALSE` draws the full query names instead.
- `plotHypeREnrichment()` draws one geneset: the running enrichment score with
  its hits and leading edge for kstest and fgsea, or a Venn diagram of the query
  and the geneset for hypergeometric results. fgsea tests the whole ranking in
  one run and only files each geneset under `| up` or `| down` by the sign of
  its enrichment score, so pass the signature name as `query` (the side holding
  the geneset is used) and the plot names the whole signature. A kstest
  `| down` query ranks the negated scores, so it keeps its side.
- `plotHypeRMap()` draws hypeR's enrichment map (genesets linked by shared
  genes). When nothing passes the cutoffs, or no pair of genesets reaches
  `similarity_cutoff`, it returns `NULL` with a warning instead of an error.

```{r, fig.width = 7, fig.height = 4.5, eval = requireNamespace("hypeR", quietly = TRUE) && requireNamespace("msigdbr", quietly = TRUE) && requireNamespace("fgsea", quietly = TRUE)}
top_up <- gsea$data[["LLFS_Aging_Gene_2023 | up"]]$data$label[1]
SigRepo::plotHypeREnrichment(gsea, top_up, query = "LLFS_Aging_Gene_2023")
```

```{r, fig.width = 6, fig.height = 4.5}
SigRepo::plotHypeREnrichment(hyp, hyp$data[[1]]$data$label[1], query = names(hyp$data)[1])
```

```{r}
SigRepo::plotHypeRMap(hyp, query = names(hyp$data)[1], top = 20, similarity_cutoff = 0.1)
```

`hypeRDotData()` and `hypeREnrichmentData()` return the numbers behind the
first two figures, for your own plots. hypeR's `hyp_dots()`, `hyp_emap()` and
`hyp_to_rmd()` still work on these results, but in hypeR 2.0.0 a single-query
`hyp_dots()` mislabels its significance axis under ggplot2 4, `merge = TRUE`
fails on a one-query result, and `hyp_emap()` errors when no geneset pair
reaches the similarity cutoff.

## Direction types

hypeR itself has no notion of direction; the split into queries is SigRepo's.

| `direction_type` | Hypergeometric (`split = TRUE`) | kstest and fgsea |
|---|---|---|
| uni-directional | one query of every feature | the whole difexp, ranked by score |
| bi-directional | one query per `group_label` (each arm is a direction) | the whole difexp, ranked by signed score; `direction` picks the end (kstest) or the ES side (fgsea) |
| categorical | one query per category and score sign: `"<label> \| <group> \| up"` / `"\| down"` (by category only when there are no scores) | each category's own difexp rows ranked separately: `"<label> \| <group>"`, with `" \| up"` / `" \| down"` for `direction = "both"` |

A categorical signature's groups are categories (e.g. subtypes), each with its
own up- and down-regulated genes, which is why its categories are split by
sign and ranked separately rather than pooled into one list.

## Background

`background = NULL`, the default, uses each signature's measured genes from its
difexp for the hypergeometric test, following BS831's definition of the
background as the annotated genes of the dataset the signature came from. It
falls back to `23467` when a signature has no difexp, and, with a warning, when
the difexp looks filtered rather than complete: a transcriptomics difexp with
fewer than 10,000 rows, a difexp with no more than twice its signature's rows,
or one where every `p_value`, `pvalue` or `adj_p` is at most 0.05. A kstest
keeps `23467`, because its ranked list already is the universe.

`background` also takes what hypeR takes, a population size or a gene vector,
plus `"difexp"`, which uses the difexp even when it looks filtered. Anything
else is an error. With a gene vector or `"difexp"`, a hypergeometric query is
first reduced to the background genes (hypeR reduces only the genesets), with a
warning naming each query that lost genes. `SigRepo Background Source` and
`SigRepo Query Genes Removed` in each `hyp$info` record what happened.

```{r}
hyp_measured <- SigRepo::runHypeR(
  omic_signature = sig,
  genesets = hallmark,
  background = "difexp",
  verbose = FALSE
)
hyp_measured$data[[1]]$info$Background
```

## Export

`hypeR::hyp_to_excel()` uses each query name as a sheet name, and Excel allows
at most 31 characters with none of `: \ / ? * [ ]`. Most SigRepo signature
names are longer, so use `hypeRToExcel()`: it shortens and de-duplicates the
sheet names and adds an `index` sheet that maps each sheet back to its query
and signature.

```{r, eval = requireNamespace("hypeR", quietly = TRUE) && requireNamespace("msigdbr", quietly = TRUE) && requireNamespace("openxlsx", quietly = TRUE)}
xlsx <- tempfile(fileext = ".xlsx")
SigRepo::hypeRToExcel(hyp, file_path = xlsx)
```

## Your own cutoffs and gene lists

`runHypeR()` builds hypergeometric queries from the stored signature table. To
use a different threshold, get the difexp with the gene symbols `runHypeR()`
resolves, filter it, and pass the genes as `signature`, which takes anything
`hypeR::hypeR()` takes: a character vector, a named numeric vector, or a named
list of those.

```{r}
difexp <- SigRepo::getHypeRDifexp(omic_signature = sig, verbose = FALSE)[[1]]
strict <- difexp[!is.na(difexp$resolved_symbol) & difexp$adj_p < 0.01, ]

custom <- SigRepo::runHypeR(
  signature = list(
    strict_up = strict$resolved_symbol[strict$score > 0],
    strict_down = strict$resolved_symbol[strict$score < 0]
  ),
  genesets = hallmark,
  background = unique(stats::na.omit(difexp$resolved_symbol)),
  verbose = FALSE
)
names(custom$data)
```

`signature` cannot be mixed with `omic_signature`, `signature_id` or
`signature_name`, and its results carry `Symbol Source = "supplied"` with no
signature ID.

## Per-signature backgrounds

`background` also takes a named list, one entry per signature, keyed by
signature label (the `signature_name` column of `prepareHypeRSignatures()$info`)
or signature ID. Each entry is a number, a gene vector, or `"difexp"`, and the
arms of a split signature share their signature's entry:

```{r, eval = FALSE}
hyp <- SigRepo::runHypeR(
  omic_signature = list(aging = sig, platform_study = other_sig),
  genesets = hallmark,
  background = list(aging = "difexp", platform_study = platform_genes)
)
```

## Signatures from the database

With a connection, pass signature ids or names instead of objects. Signatures
that cannot produce a query (no gene symbols, no difexp for `kstest`) are
skipped with a warning.

```{r, eval = FALSE}
conn_handler <- SigRepo::newConnHandler(
  dbname = "sigrepo",
  host = "sigrepo.org",
  port = 3306,
  user = "<user-name>",
  password = "<password>"
)

hyp <- SigRepo::runHypeR(
  conn_handler = conn_handler,
  signature_id = c(101, 102, 103),
  genesets = hallmark,
  fdr = 0.05
)
SigRepo::hypeRToExcel(hyp, file_path = "sigrepo_hallmark.xlsx")
```
