---
title: "Running hypeR enrichment on SigRepo signatures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running hypeR enrichment on SigRepo signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  eval = requireNamespace("hypeR", quietly = TRUE) && requireNamespace("msigdbr", quietly = TRUE)
)
```

```{r}
library(SigRepo)
library(hypeR)

utils::data("LLFS_Aging_Gene_2023", package = "SigRepo")
sig <- LLFS_Aging_Gene_2023
```

## What `runHypeR()` does

`runHypeR()` turns SigRepo signatures into the query vectors `hypeR::hypeR()`
expects, runs hypeR, and returns hypeR's own object. Every argument from `test`
on is hypeR's, with hypeR's defaults, so a SigRepo run and a plain hypeR run on
the same genes give the same table.

## Genesets

Fetch an MSigDB collection with `genesets = "msigdb"`:

```{r}
hallmark <- SigRepo::getHypeRGenesets("msigdb", msigdb_collection = "H")
hallmark
```

Or pass your own genesets — a named list, `hypeR::gsets` or `hypeR::rgsets`:

```{r}
my_genesets <- list(
  ribosome_like = c("RPL3", "RPL4", "RPS6"),
  interferon_like = c("ISG15", "IFIT1", "MX1")
)
```

`runHypeR()` accepts either form in `genesets`. The `msigdb_*` arguments only
apply with `genesets = "msigdb"`.

## Inspect the query vectors

`prepareHypeRSignatures()` shows exactly what will be tested and where the gene
symbols came from:

```{r}
prepared <- SigRepo::prepareHypeRSignatures(omic_signature = sig, verbose = FALSE)
prepared$info
lengths(prepared$signatures)
```

This signature's table has no symbol column, so symbols come from its difexp,
joined on `probe_id`. A bi-directional signature has one `group_label` per arm,
and `split = TRUE` (the default) tests each arm separately. Query names are
always unique — built as `"<label> | <group_label>"` (or just `"<label>"` when
there is no group) and disambiguated as `name (2)`, `name (3)` if two
signatures would otherwise collide.

## Hypergeometric test

```{r}
hyp <- SigRepo::runHypeR(
  omic_signature = sig,
  genesets = "msigdb",
  msigdb_collection = "H",
  verbose = FALSE
)
hyp
```

As in hypeR, `fdr` and `pval` default to `1`, so every geneset is kept; pass
`fdr = 0.05` to filter. The result is a hypeR `multihyp` (one `hyp` per arm),
so hypeR's functions work on it directly:

```{r, fig.width = 7, fig.height = 5}
hypeR::hyp_dots(hyp, merge = TRUE, top = 10, title = "LLFS aging, Hallmark")
```

Each `hyp$info` records where its query came from:

```{r}
tail(hyp$data[[1]]$info, 4)
```

## Ranked (KS) test

`test = "kstest"` ranks the full difexp table by `score`. `power = 1` (hypeR's
default) weights hits by score, GSEA-style; `power = 0` is the classic KS test.
With `power != 0`, a geneset whose hits in a query all score exactly `0` cannot
be scored by a weighted KS test, so it is dropped for that query with a
warning.

```{r}
ranked <- SigRepo::runHypeR(
  omic_signature = sig,
  genesets = hallmark,
  test = "kstest",
  power = 0,
  verbose = FALSE
)
head(ranked$data)
```

## Background

`background` takes what hypeR takes — a population size (default `23467`) or a
gene vector — plus `"difexp"`, which uses each signature's measured genes.
Anything else is an error: `background` must be a number, a gene vector, or
exactly `"difexp"`.

```{r}
hyp_measured <- SigRepo::runHypeR(
  omic_signature = sig,
  genesets = hallmark,
  background = "difexp",
  verbose = FALSE
)
hyp_measured$data[[1]]$info$Background
```

## Signatures from the database

With a connection, pass signature ids or names instead of objects. Signatures
that cannot produce a query (no gene symbols, no difexp for `kstest`) are
skipped with a warning.

```{r, eval = FALSE}
conn_handler <- SigRepo::newConnHandler(
  dbname = "sigrepo",
  host = "sigrepo.org",
  port = 3306,
  user = "<user-name>",
  password = "<password>"
)

hyp <- SigRepo::runHypeR(
  conn_handler = conn_handler,
  signature_id = c(101, 102, 103),
  genesets = "msigdb",
  msigdb_collection = "H",
  fdr = 0.05
)

# hyp_to_excel() uses each query name as an Excel sheet name, which must be at
# most 31 characters and cannot contain : \ / ? * [ ], and SigRepo query names
# built from long signature names routinely break that, so make them sheet-safe first:
names(hyp$data) <- make.unique(substr(gsub("[][\\\\/?*:]", "_", names(hyp$data)), 1, 28))
hypeR::hyp_to_excel(hyp, file_path = "sigrepo_hallmark.xlsx")
```
