---
title: "Running hypeR enrichment on SigRepo signatures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running hypeR enrichment on SigRepo signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  eval = requireNamespace("hypeR", quietly = TRUE) && requireNamespace("msigdbr", quietly = TRUE)
)
```

```{r}
library(SigRepo)
library(hypeR)

utils::data("LLFS_Aging_Gene_2023", package = "SigRepo")
sig <- LLFS_Aging_Gene_2023
```

## What `runHypeR()` does

`runHypeR()` turns SigRepo signatures into the query vectors `hypeR::hypeR()`
expects, runs hypeR, and returns hypeR's own object. Every argument from
`background` on is hypeR's, with hypeR's defaults, so a SigRepo run and a plain
hypeR run on the same genes give the same table.

The hypeR documentation site (montilab.github.io/hypeR-docs) still shows hypeR
1.7.2. hypeR 2.x renamed some helper arguments, e.g.
`msigdb_gsets(category =, subcategory =)` is now `collection =` /
`subcollection =`. SigRepo needs hypeR 2.0.0 or later, so check the help pages
of your installed hypeR (`?hypeR::msigdb_gsets`) rather than the site.

## Genesets

Fetch an MSigDB collection with `genesets = "msigdb"`:

```{r}
hallmark <- SigRepo::getHypeRGenesets("msigdb", msigdb_collection = "H")
hallmark
```

Or pass your own genesets — a named list, `hypeR::gsets` or `hypeR::rgsets`:

```{r}
my_genesets <- list(
  ribosome_like = c("RPL3", "RPL4", "RPS6"),
  interferon_like = c("ISG15", "IFIT1", "MX1")
)
```

`runHypeR()` accepts either form in `genesets`. The `msigdb_*` arguments only
apply with `genesets = "msigdb"`. `genesets = "msigdb"` queries msigdbr on every
call, so when you run several analyses, fetch the collection once with
`getHypeRGenesets()` and pass the object, as the examples below do. Enrichr
libraries (`hypeR::enrichr_gsets()`) and hypeR's curated hierarchies
(`hypeR::hyperdb_rgsets()`, used by `hyp_hmap()`) pass through the same way.

## Inspect the query vectors

`prepareHypeRSignatures()` shows exactly what will be tested and where the gene
symbols came from:

```{r}
prepared <- SigRepo::prepareHypeRSignatures(omic_signature = sig, verbose = FALSE)
prepared$info
lengths(prepared$signatures)
```

This signature's table has no symbol column, so symbols come from its difexp,
joined on `probe_id`. A bi-directional signature has one `group_label` per arm,
and `split = TRUE` (the default) tests each arm separately. Query names are
always unique — built as `"<label> | <group_label>"` (or just `"<label>"` when
there is no group) and disambiguated as `name (2)`, `name (3)` if two
signatures would otherwise collide. To choose your own names, pass
`query_names`, a function of the `info` table:

```{r}
short <- SigRepo::prepareHypeRSignatures(
  omic_signature = sig,
  query_names = function(info) paste0("LLFS_", info$group_label),
  verbose = FALSE
)
names(short$signatures)
```

## Hypergeometric test

```{r}
hyp <- SigRepo::runHypeR(
  omic_signature = sig,
  genesets = hallmark,
  verbose = FALSE
)
hyp
```

As in hypeR, `fdr` and `pval` default to `1`, so every geneset is kept; pass
`fdr = 0.05` to filter. The result is a hypeR `multihyp` (one `hyp` per arm),
so hypeR's functions work on it directly:

```{r, fig.width = 7, fig.height = 5}
hypeR::hyp_dots(hyp, merge = TRUE, top = 10, title = "LLFS aging, Hallmark")
```

`hyp_show()` takes a single `hyp`, so pass one arm:

```{r, eval = FALSE}
hypeR::hyp_show(hyp$data[[1]])
```

The return type follows the input, as in hypeR. A single signature (one
`OmicSignature`, id or name) gives a `hyp` when it can only build one query:
`split = FALSE`, or a kstest with `direction = "up"` or `"down"`. Everything
else gives a `multihyp`, even when only one query is left after skips, so
batch code can always index `res$data`.

Each `hyp$info` records where its query came from and how it was run:

```{r}
tail(hyp$data[[1]]$info, 13)
```

## Ranked (KS) test

`test = "kstest"` ranks the full difexp table by `score`, highest first.

- The p-value and FDR come from hypeR's unweighted, one-sided KS test. They do
  not depend on `power` or `absolute`.
- `power` changes only the enrichment `score`: `1` (hypeR's default) weights
  hits by |score|; `0` runs hypeR's ranked signature (gene names in rank
  order), so `Signature Type` is `"ranked"`.
- The test finds genesets enriched toward the top of the ranking only.
  `direction = "down"` ranks by the negated score, so genesets concentrated
  among the most negative scores come first; `direction = "both"` runs both and
  returns `"<label> | up"` and `"<label> | down"`.
- `ks_source = "signature"` ranks the signature table instead of the difexp,
  for signatures stored without one. Only the signature's genes are ranked, so
  the test asks where genesets fall within the signature, not within
  everything measured.

hypeR's kstest cannot score a geneset whose hits in a query all score exactly
`0` (with `power != 0`), or one that contains every query gene (at any power).
Such genesets are dropped for that query with a warning, which also leaves them
out of the FDR adjustment. A query with no genesets left is skipped with a
warning.

```{r}
ranked <- SigRepo::runHypeR(
  omic_signature = sig,
  genesets = hallmark,
  test = "kstest",
  direction = "both",
  verbose = FALSE
)
names(ranked$data)
head(ranked$data[["LLFS_Aging_Gene_2023 | down"]]$data)
```

## Background

`background` takes what hypeR takes — a population size (default `23467`) or a
gene vector — plus `"difexp"`, which uses each signature's measured genes.
Anything else is an error: `background` must be a number, a gene vector, or
exactly `"difexp"`. With a gene vector or `"difexp"`, a hypergeometric query is
first reduced to the background genes (hypeR reduces only the genesets), with a
warning naming each query that lost genes. `SigRepo Background Source` and
`SigRepo Query Genes Removed` in each `hyp$info` record what happened.

```{r}
hyp_measured <- SigRepo::runHypeR(
  omic_signature = sig,
  genesets = hallmark,
  background = "difexp",
  verbose = FALSE
)
hyp_measured$data[[1]]$info$Background
```

## Export

`hypeR::hyp_to_excel()` uses each query name as a sheet name, and Excel allows
at most 31 characters with none of `: \ / ? * [ ]`. Most SigRepo signature
names are longer, so use `hypeRToExcel()`: it shortens and de-duplicates the
sheet names and adds an `index` sheet that maps each sheet back to its query
and signature.

```{r, eval = requireNamespace("hypeR", quietly = TRUE) && requireNamespace("msigdbr", quietly = TRUE) && requireNamespace("openxlsx", quietly = TRUE)}
xlsx <- tempfile(fileext = ".xlsx")
SigRepo::hypeRToExcel(hyp, file_path = xlsx)
```

## Signatures from the database

With a connection, pass signature ids or names instead of objects. Signatures
that cannot produce a query (no gene symbols, no difexp for `kstest`) are
skipped with a warning.

```{r, eval = FALSE}
conn_handler <- SigRepo::newConnHandler(
  dbname = "sigrepo",
  host = "sigrepo.org",
  port = 3306,
  user = "<user-name>",
  password = "<password>"
)

hyp <- SigRepo::runHypeR(
  conn_handler = conn_handler,
  signature_id = c(101, 102, 103),
  genesets = hallmark,
  fdr = 0.05
)
SigRepo::hypeRToExcel(hyp, file_path = "sigrepo_hallmark.xlsx")
```
