---
title: "Derivation of signatures from the Long Life Family Study (LLFS) plasma metabolome"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
    toc_float: true
---

```{r global, echo=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

```{r}
current_input <- knitr::current_input(
  dir = TRUE
)

if (nzchar(current_input)) {
  RMD_DIR <- dirname(
    normalizePath(
      current_input,
      winslash = "/",
      mustWork = TRUE
    )
  )
} else {
  RMD_DIR <- normalizePath(
    getwd(),
    winslash = "/",
    mustWork = TRUE
  )
}

REFERENCE_DIR <- file.path(
  RMD_DIR,
  "reference_data"
)

OUTDIR <- file.path(
  RMD_DIR,
  "outputs"
)

INFILE <- Sys.getenv(
  "LLFS_INPUT_FILE",
  unset = ""
)

if (!nzchar(INFILE)) {
  stop(
    paste0(
      "LLFS_INPUT_FILE is not set. ",
      "Set it to the LLFS supplementary ",
      "Excel workbook path before rendering."
    ),
    call. = FALSE
  )
}

stopifnot(
  file.exists(INFILE),
  dir.exists(REFERENCE_DIR)
)

## OmicSignature tutorial
## montilab.github.io/OmicSignature/articles/CreateOmS.html

## support scripts
## check github.com/montilab/MLscripts for useful scripts
## e.g.,
## > options(box.path=file.path(Sys.getenv("CBMGIT"), "MLscripts"))
## > box::use(R/deseq_wrapper)
## alternative (less desirable)
## > source(file.path(Sys.getenv("CBMGIT"), "MLscripts", "R", "deseq_wrapper.R"))
```

```{r}
user_lib <- Sys.getenv(
  "SIGREPO_R_LIB",
  unset = ""
)

if (nzchar(user_lib)) {
  stopifnot(
    dir.exists(user_lib)
  )

  .libPaths(
    c(
      user_lib,
      .libPaths()
    )
  )
}

library(OmicSignature)
library(dplyr)
library(stringr)
library(openxlsx)
library(readxl)
library(purrr)
library(tibble)

library(tidyr)

library(biomaRt)

library(curl)
library(stringi)


find_codex_reference_root <- function(start = getwd()) {
  current <- normalizePath(
    start,
    winslash = "/",
    mustWork = TRUE
  )

  repeat {
    candidate <- file.path(current, "codex_reference")

    if (dir.exists(candidate)) {
      return(
        normalizePath(
          candidate,
          winslash = "/",
          mustWork = TRUE
        )
      )
    }

    parent <- dirname(current)

    if (identical(parent, current)) {
      stop(
        "Could not locate the codex_reference directory.",
        call. = FALSE
      )
    }

    current <- parent
  }
}

CODEX_REFERENCE_ROOT <- find_codex_reference_root(
  RMD_DIR
)

OMIC_HELPER <- file.path(
  CODEX_REFERENCE_ROOT,
  "helpers",
  "omicsignature_compat.R"
)

stopifnot(file.exists(OMIC_HELPER))
source(OMIC_HELPER)

OMIC_API <- sigrepo_assert_compatible_api()
print(OMIC_API)

```

# Preflight interpretation

## Study and source

- Study: Sebastiani et al., Cell Reports, 2024.
- PMID: 39504246.
- Source workbook sheet: `all results-regression`.
- Platform: metabolomics by mass spectrometry.
- Sample: blood plasma.
- The paper reports RefMet standardized names from the July 2023 release.

## Biologically justified objects

1. Longitudinal aging-associated plasma metabolites.
2. Exceptional-longevity-associated plasma metabolites.
3. Survival-associated plasma metabolites.
4. One collection containing exactly those three related objects.

## Source classification and `difexp`

The workbook contains a broad regression-results table with effect estimates,
adjusted P values, and study-provided Boolean selection indicators. The mapped
parent result table is retained as `difexp`; each signature is derived from the
same parent table using its corresponding study selection indicator.

## Score and direction

- Longitudinal aging: the reported aging effect.
- Exceptional longevity: the reported centenarian effect.
- Survival: `log(HR.death.est)`.
- Zero-valued effects are not assigned to either biological direction.
- Group labels are factors with explicit levels.

## Feature and probe identifiers

- `Standardized.name` is the primary RefMet `feature_name`.
- `refmet_convert()` is rescue-only when `Standardized.name` is missing.
- `Compound.Name` is used as the source feature/probe identifier only after
  its non-missingness and uniqueness are validated.
- Rows unresolved after RefMet rescue remain in the mapping audit and are
  excluded only from object input.

# Source manifest

| Object family | Source sheet | Parent table | Selection column | `difexp` |
|---|---|---|---|---|
| Longitudinal aging | all results-regression | complete mapped regression results | aging.fdr.05 | retained |
| Exceptional longevity | all results-regression | complete mapped regression results | el.fdr.05 | retained |
| Survival | all results-regression | complete mapped regression results | surv.fdr.05 | retained |




It includes the star-cleaning step, RefMet rescue, and QC tables.
```{r}
refmet_convert <- function(DF) {
  stopifnot(methods::is(DF, "data.frame"))

  mets <- stringi::stri_join_list(list(DF |> dplyr::pull(1)), sep = "\n")
  h <- curl::new_handle()
  curl::handle_setform(h, metabolite_name = mets)

  req <- curl::curl_fetch_memory(
    "https://www.metabolomicsworkbench.org/databases/refmet/name_to_refmet_new_min.php",
    handle = h
  )

  x <- rawToChar(req$content)
  y <- strsplit(x, "\n")
  refmet <- data.frame(ncol = 7)

  for (i in seq_along(y[[1]])) {
    if (nchar(y[[1]][i]) > 1) {
      z <- strsplit(y[[1]][i], "\t")
      for (j in seq_along(z[[1]])) {
        refmet[i, j] <- z[[1]][j]
      }
    }
  }

  refmet <- refmet[rowSums(is.na(refmet)) != ncol(refmet), , drop = FALSE]
  colnames(refmet) <- refmet[1, ]
  refmet <- refmet[-1, ]
  refmet[is.na(refmet)] <- ""

  refmet |>
    dplyr::rename_with(~ trimws(.x)) |>
    dplyr::rename_with(~ tolower(stringr::str_replace(.x, "[ ]+", "_")), everything()) |>
    dplyr::rename(
      metabolite = "input_name",
      refmet_name = "standardized_name"
    ) |>
    dplyr::mutate(refmet_name = ifelse(refmet_name == "-", NA, refmet_name)) |>
    dplyr::mutate(across(-c(metabolite, refmet_name), ~ ifelse(.x == "", NA, .x)))
}
```

```{r}
clean_text <- function(x) {
  x <- as.character(x)
  stringr::str_squish(x)
}

clean_compound_for_refmet <- function(x) {
  x <- clean_text(x)
  x <- stringr::str_replace(x, "\\*$", "")   # remove trailing star only
  x <- stringr::str_squish(x)
  x
}

metab_raw0 <- readxl::read_excel(
  path = INFILE,
  sheet = "all results-regression"
) %>%
  dplyr::mutate(
    source_file = basename(INFILE),
    source_sheet = "all results-regression",
    source_row = dplyr::row_number() + 1L,
    Compound.Name = clean_text(Compound.Name),
    Compound.Name_clean = clean_compound_for_refmet(Compound.Name),
    Standardized.name = clean_text(Standardized.name),
    Standardized.name = stringr::str_replace(Standardized.name, "\\*$", ""),
    Standardized.name = clean_text(Standardized.name),
    Standardized.name = dplyr::if_else(
     is.na(Standardized.name) | Standardized.name == "" | Standardized.name == "-",
     NA_character_,
     Standardized.name
)
  )
```

```{r}
# This verifies the assumption before using Compound.Name as probe_id
compound_probe_qc <- metab_raw0 |>
  dplyr::filter(!is.na(Compound.Name), nzchar(Compound.Name)) |>
  dplyr::count(Compound.Name, name = "source_row_count") |>
  dplyr::filter(source_row_count > 1L)

stopifnot(
  all(!is.na(metab_raw0$Compound.Name)),
  all(nzchar(metab_raw0$Compound.Name)),
  nrow(compound_probe_qc) == 0L
)
```

```{r refmet-rescue}
## RefMet rescue is cache-first for reproducibility.
## Generate the cache once in a controlled environment, review it, and keep it
## with the approved reference materials.

REFMET_CACHE <- file.path(
  REFERENCE_DIR,
  "LLFS_Sebastiani2024_refmet_rescue_2023_07.rds"
)

stopifnot(
  file.exists(REFMET_CACHE)
)

ALLOW_LIVE_REFMET_QUERY <- FALSE                                

compound_df <- metab_raw0 |>
  dplyr::distinct(Compound.Name_clean) |>
  dplyr::filter(!is.na(Compound.Name_clean), nzchar(Compound.Name_clean)) |>
  dplyr::rename(metabolite = Compound.Name_clean)

if (file.exists(REFMET_CACHE)) {
  refmet_rescue <- readRDS(REFMET_CACHE)
} else if (ALLOW_LIVE_REFMET_QUERY) {
  refmet_rescue <- refmet_convert(compound_df) |>
    dplyr::transmute(
      Compound.Name_clean = clean_text(metabolite),
      refmet_name_rescued = clean_text(refmet_name),
      resolver = "Metabolomics Workbench RefMet name converter",
      retrieved_at_utc = format(Sys.time(), tz = "UTC", usetz = TRUE)
    ) |>
    dplyr::mutate(
      refmet_name_rescued = dplyr::if_else(
        is.na(refmet_name_rescued) |
          !nzchar(refmet_name_rescued) |
          refmet_name_rescued == "-",
        NA_character_,
        refmet_name_rescued
      )
    ) |>
    dplyr::distinct()

  stop(
    paste0(
      "A new RefMet rescue table was generated in memory. Review it and save ",
      "the approved cache explicitly as: ",
      REFMET_CACHE,
      ". Then rerun with ALLOW_LIVE_REFMET_QUERY = FALSE."
    ),
    call. = FALSE
  )
} else {
  stop(
    paste0(
      "Required reviewed RefMet rescue cache is missing: ",
      REFMET_CACHE,
      ". Do not query a live resolver silently in an approved workflow."
    ),
    call. = FALSE
  )
}

stopifnot(
  is.data.frame(refmet_rescue),
  all(c("Compound.Name_clean", "refmet_name_rescued") %in%
        names(refmet_rescue)),
  !anyDuplicated(refmet_rescue$Compound.Name_clean)
)
```

```{r}
## Final feature_name precedence:
## 1) paper-provided Standardized.name
## 2) rescued RefMet name from Compound.Name
## 3) NA -> retained in audit and excluded from object input

metab_raw <- metab_raw0 |>
  dplyr::left_join(refmet_rescue, by = "Compound.Name_clean") |>
  dplyr::mutate(
    ## Compound.Name is a unique source-table feature identifier in this workbook.
    probe_id = as.character(Compound.Name),
    feature_name = dplyr::coalesce(Standardized.name, refmet_name_rescued),
    feature_name = stringr::str_replace(feature_name, "\\*$", ""),
    feature_name = clean_text(feature_name),
    refmet_source = dplyr::case_when(
      !is.na(Standardized.name) ~ "provided_Standardized.name",
      is.na(Standardized.name) & !is.na(refmet_name_rescued) ~
        "rescued_from_Compound.Name_clean",
      TRUE ~ "missing_refmet"
    )
  )
```

```{r}
qc_source_summary <- dplyr::bind_rows(
  metab_raw %>%
    dplyr::filter(aging.fdr.05 == TRUE) %>%
    dplyr::count(refmet_source) %>%
    dplyr::mutate(signature = "Aging"),

  metab_raw %>%
    dplyr::filter(el.fdr.05 == TRUE) %>%
    dplyr::count(refmet_source) %>%
    dplyr::mutate(signature = "EL"),

  metab_raw %>%
    dplyr::filter(surv.fdr.05 == TRUE) %>%
    dplyr::count(refmet_source) %>%
    dplyr::mutate(signature = "Survival")
) %>%
  dplyr::select(signature, refmet_source, n)

print(qc_source_summary)
```

```{r}
qc_summary <- tibble::tibble(
  signature = c("Aging", "EL", "Survival"),

  original_significant = c(
    sum(metab_raw$aging.fdr.05 == TRUE, na.rm = TRUE),
    sum(metab_raw$el.fdr.05 == TRUE, na.rm = TRUE),
    sum(metab_raw$surv.fdr.05 == TRUE, na.rm = TRUE)
  ),

  original_refmet_from_paper = c(
    sum(metab_raw$aging.fdr.05 == TRUE & !is.na(metab_raw$Standardized.name), na.rm = TRUE),
    sum(metab_raw$el.fdr.05 == TRUE & !is.na(metab_raw$Standardized.name), na.rm = TRUE),
    sum(metab_raw$surv.fdr.05 == TRUE & !is.na(metab_raw$Standardized.name), na.rm = TRUE)
  ),

  rescued_new_refmet = c(
    sum(metab_raw$aging.fdr.05 == TRUE &
          metab_raw$refmet_source == "rescued_from_Compound.Name_clean", na.rm = TRUE),
    sum(metab_raw$el.fdr.05 == TRUE &
          metab_raw$refmet_source == "rescued_from_Compound.Name_clean", na.rm = TRUE),
    sum(metab_raw$surv.fdr.05 == TRUE &
          metab_raw$refmet_source == "rescued_from_Compound.Name_clean", na.rm = TRUE)
  ),

  final_usable_refmet = c(
    sum(metab_raw$aging.fdr.05 == TRUE & !is.na(metab_raw$feature_name), na.rm = TRUE),
    sum(metab_raw$el.fdr.05 == TRUE & !is.na(metab_raw$feature_name), na.rm = TRUE),
    sum(metab_raw$surv.fdr.05 == TRUE & !is.na(metab_raw$feature_name), na.rm = TRUE)
  ),

  still_unmapped = c(
    sum(metab_raw$aging.fdr.05 == TRUE & is.na(metab_raw$feature_name), na.rm = TRUE),
    sum(metab_raw$el.fdr.05 == TRUE & is.na(metab_raw$feature_name), na.rm = TRUE),
    sum(metab_raw$surv.fdr.05 == TRUE & is.na(metab_raw$feature_name), na.rm = TRUE)
  )
)

print(qc_summary)
```

```{r threshold-validation}
threshold_flag_qc <- tibble::tibble(
  signature = c(
    "Longitudinal aging",
    "Exceptional longevity",
    "Survival"
  ),
  mismatch_n = c(
    sum(
      dplyr::coalesce(
        as.logical(metab_raw$aging.fdr.05),
        FALSE
      ) !=
        (
          !is.na(metab_raw$Aging_pval_adj) &
            metab_raw$Aging_pval_adj < 0.05
        )
    ),
    sum(
      dplyr::coalesce(
        as.logical(metab_raw$el.fdr.05),
        FALSE
      ) !=
        (
          !is.na(metab_raw$EL_pval_adj) &
            metab_raw$EL_pval_adj < 0.05
        )
    ),
    sum(
      dplyr::coalesce(
        as.logical(metab_raw$surv.fdr.05),
        FALSE
      ) !=
        (
          !is.na(metab_raw$Metab_pval_adj) &
            metab_raw$Metab_pval_adj < 0.05
        )
    )
  )
)

print(threshold_flag_qc)

stopifnot(
  all(threshold_flag_qc$mismatch_n == 0L)
)
```

 - Use the original Metabolon compound names (Compound.Name)
 
 - Re-query the current RefMet database
 
 - Recover additional mappings because:
 
   * RefMet database evolves over time
   
   * newer RefMet versions may now recognize compounds that were previously unmapped
   
   * formatting issues (stars, suffixes, spaces) may prevent matching

```{r}
dir.create(
  OUTDIR,
  recursive = TRUE,
  showWarnings = FALSE
)

do_write_audit <- FALSE

if (do_write_audit) {
  saveRDS(
    metab_raw,
    file = file.path(
      OUTDIR,
      "LLFS_Metabolomics_Aging_EL_Survival_refmet_rescued_prepared.rds"
    )
  )

  audit_file <- file.path(
    OUTDIR,
    "LLFS_Metabolomics_Aging_EL_Survival_refmet_rescue_audit.xlsx"
  )

  stopifnot(!file.exists(audit_file))

  openxlsx::write.xlsx(
    list(
      prepared_table = as.data.frame(metab_raw),
      refmet_rescue_table = as.data.frame(refmet_rescue),
      qc_source_summary = as.data.frame(qc_source_summary),
      qc_summary = as.data.frame(qc_summary)
    ),
    file = audit_file,
    overwrite = FALSE
  )
}

## saves the fully processed metabolomics table after:
# cleaning, rescue mapping, feature_name assignment and QC annotations

## Contains:
# original compound names, original standardized names, rescued RefMet names,
# final feature_name, source of mapping, all statistics/effects/FDR columns
```


# 2) OmicSignature object
Continue using the in-memory `metab_raw` object.

```{r}
AUTHOR_STR <- "Sebastiani2024"
AUTHOR <- "Sebastiani P et al."

ORGANISM_STR <- "Homo sapiens"
ASSAY_STR <- "metabolomics"
PLATFORM_STR <- "metabolomics by mass spectrometry"
SAMPLE_STR <- "blood plasma"

sigrepo_assert_platform(PLATFORM_STR)
sigrepo_assert_sample_type(SAMPLE_STR)

YEAR_NUM <- 2024
PMID_STR <- "39504246"

KEYWORDS <- c(
  "LLFS",
  "Long Life Family Study",
  "metabolomics",
  "Metabolon",
  "aging",
  "exceptional longevity",
  "survival"
)
```

```{r}
make_safe_name <- function(x) {
  x %>%
    as.character() %>%
    str_replace_all("[^A-Za-z0-9]+", "_") %>%
    str_replace_all("_+", "_") %>%
    str_replace_all("^_|_$", "")
}

sanitize_for_sql <- function(x) {
  x <- as.character(x)
  x <- gsub("[\r\n\t]+", " ", x)
  x <- gsub("'", "", x, fixed = TRUE)
  str_squish(x)
}

make_signature_name <- function(detail) {
  paste(
    make_safe_name("Aging"),
    make_safe_name("Hs"),
    make_safe_name("BloodPlasma"),
    make_safe_name(detail),
    make_safe_name(AUTHOR_STR),
    sep = "_"
  )
}

make_collection_name <- function(topic) {
  paste(
    make_safe_name("Aging"),
    make_safe_name("Hs"),
    make_safe_name("BloodPlasma"),
    make_safe_name(topic),
    make_safe_name(AUTHOR_STR),
    sep = "_"
  )
}
```

```{r}
make_metadata <- function(signature_name,
                          phenotype,
                          description,
                          covariates,
                          keywords) {
  OmicSignature::createMetadata(
    signature_name = sanitize_for_sql(signature_name),
    organism = ORGANISM_STR,
    direction_type = "bi-directional",
    assay_type = ASSAY_STR,
    phenotype = sanitize_for_sql(phenotype),
    covariates = covariates,
    platform = PLATFORM_STR,
    sample_type = SAMPLE_STR,
    author = AUTHOR,
    year = YEAR_NUM,
    PMID = PMID_STR,
    keywords = sigrepo_keywords(keywords),
    adj_p_cutoff = 0.05,
    description = sanitize_for_sql(description)
  )
}
```
Important changes:
 - actual covariates are supplied per object;
 - keyword terms are passed through `sigrepo_keywords()` and stored as one
   scalar metadata string, as required by the validated OmicSignature API contract;
 - `score_cutoff = 0` is removed because zero was not a study selection threshold.


```{r}
prepare_metab_parent <- function(df,
                                 score_col,
                                 adjp_col,
                                 label_pos,
                                 label_neg,
                                 survival = FALSE) {
  df |>
    dplyr::mutate(
      score = if (survival) {
        log(as.numeric(.data[[score_col]]))
      } else {
        as.numeric(.data[[score_col]])
      },
      adj_p = as.numeric(.data[[adjp_col]]),
      group_label = factor(
        dplyr::case_when(
          score > 0 ~ label_pos,
          score < 0 ~ label_neg,
          TRUE ~ NA_character_
        ),
        levels = c(label_pos, label_neg)
      )
    ) |>
    dplyr::filter(
      !is.na(feature_name),
      nzchar(feature_name),
      feature_name != "-",
      !is.na(probe_id),
      nzchar(probe_id),
      !is.na(score),
      is.finite(score),
       score != 0,
      !is.na(group_label)
    ) |>
    dplyr::arrange(
      is.na(adj_p),
      adj_p,
      dplyr::desc(abs(score))
    ) |>
    dplyr::distinct(
      probe_id,
      feature_name,
      score,
      adj_p,
      group_label,
      .keep_all = TRUE
    ) |>
    as.data.frame()
}
# Rationale: rows with valid analytical effects remain in the parent difexpeven when an adjusted P value is missing. The final selected signature stillmust have a finite adjusted P value and satisfy the study selection flag.

derive_signature <- function(parent, flag_col) {
  parent |>
    dplyr::filter(.data[[flag_col]] %in% TRUE) |>
    dplyr::select(
      probe_id,
      feature_name,
      score,
      group_label
    ) |>
    as.data.frame()
}

# This creates each complete parent table once and derives the selected signature from that same parent.
```

```{r}
difexp_aging <- prepare_metab_parent(
  metab_raw,
  score_col = "Aging.effect",
  adjp_col = "Aging_pval_adj",
  label_pos = "Increasing with longitudinal age",
  label_neg = "Decreasing with longitudinal age"
)

sig_aging <- derive_signature(
  parent = difexp_aging,
  flag_col = "aging.fdr.05"
)

difexp_el <- prepare_metab_parent(
  metab_raw,
  score_col = "cent.effect",
  adjp_col = "EL_pval_adj",
  label_pos = "Higher in exceptional longevity",
  label_neg = "Lower in exceptional longevity"
)

sig_el <- derive_signature(
  parent = difexp_el,
  flag_col = "el.fdr.05"
)

difexp_survival <- prepare_metab_parent(
  metab_raw,
  score_col = "HR.death.est",
  adjp_col = "Metab_pval_adj",
  label_pos = "Higher mortality risk",
  label_neg = "Lower mortality risk",
  survival = TRUE
)

sig_survival <- derive_signature(
  parent = difexp_survival,
  flag_col = "surv.fdr.05"
)
```

```{r}
meta_aging <- make_metadata(
  signature_name = make_signature_name("LongitudinalAgingMetabolites"),
  phenotype = "Longitudinal aging-associated plasma metabolites",
  covariates = paste(
    c(
      "age at first assessment",
      "sex",
      "education",
      "enrollment country",
      "medications for type 2 diabetes",
      "dyslipidemia",
      "heart disease",
      "hypertension",
      "four genome-wide principal components"
    ),
    collapse = ", "
  ),
  keywords = c(
    "LLFS",
    "metabolomics",
    "longitudinal aging",
    "plasma"
  ),
  description = paste0(
    "Bi-directional plasma metabolomics signature of metabolites associated ",
    "with longitudinal aging in the Long Life Family Study. Positive scores ",
    "indicate metabolites increasing with longitudinal age; negative scores ",
    "indicate metabolites decreasing with longitudinal age. Feature names use ",
    "the study-provided RefMet standardized names when available, with ",
    "Compound.Name-based RefMet conversion used only as rescue."
  )
)

meta_el <- make_metadata(
  signature_name = make_signature_name("ExceptionalLongevityMetabolites"),
  phenotype = "Exceptional longevity-associated plasma metabolites",
  covariates = paste(
    c(
      "sex",
      "education",
      "enrollment country",
      "medications for type 2 diabetes",
      "dyslipidemia",
      "heart disease",
      "hypertension",
      "four genome-wide principal components"
    ),
    collapse = ", "
  ),
  keywords = c(
    "LLFS",
    "metabolomics",
    "exceptional longevity",
    "plasma"
  ),
  description = paste0(
    "Bi-directional plasma metabolomics signature of metabolites associated ",
    "with exceptional longevity in the Long Life Family Study. Positive scores ",
    "indicate metabolites higher in exceptional-longevity participants; ",
    "negative scores indicate metabolites lower in exceptional-longevity ",
    "participants. Feature names use the study-provided RefMet standardized ",
    "names when available, with Compound.Name-based RefMet conversion used ",
    "only as rescue."
  )
)

meta_survival <- make_metadata(
  signature_name = make_signature_name("SurvivalAssociatedMetabolites"),
  phenotype = "Survival-associated plasma metabolites",
  covariates = paste(
    c(
      "age at enrollment",
      "sex",
      "education",
      "enrollment country",
      "medications for type 2 diabetes",
      "dyslipidemia",
      "heart disease",
      "hypertension",
      "four genome-wide principal components"
    ),
    collapse = ", "
  ),
  keywords = c(
    "LLFS",
    "metabolomics",
    "survival",
    "mortality",
    "plasma"
  ),
  description = paste0(
    "Bi-directional plasma metabolomics signature of metabolites associated ",
    "with survival in the Long Life Family Study. Scores are log hazard ratios ",
    "for death; positive scores indicate higher mortality risk and negative ",
    "scores indicate lower mortality risk. Feature names use the study-provided ",
    "RefMet standardized names when available, with Compound.Name-based RefMet ",
    "conversion used only as rescue."
  )
)
```

Validation before object creation
```{r}
validate_metab_tables <- function(signature, difexp, flag_col) {

  stopifnot(
    all(c("probe_id", "feature_name", "score", "group_label") %in%
          names(signature)),
    all(c(
      "probe_id", "feature_name", "score", "adj_p",
      "group_label", flag_col
    ) %in% names(difexp)),
    is.character(signature$probe_id),
    is.character(difexp$probe_id),
    all(!is.na(signature$probe_id) & nzchar(signature$probe_id)),
    all(!is.na(difexp$probe_id) & nzchar(difexp$probe_id)),
    !anyDuplicated(signature$probe_id),
    !anyDuplicated(difexp$probe_id),
    all(!is.na(signature$feature_name) & nzchar(signature$feature_name)),
    is.numeric(signature$score),
    all(is.finite(signature$score)),
    all(signature$score != 0),
    is.factor(signature$group_label),
    all(!is.na(signature$group_label))
  )

  parent_lookup <- difexp |>
    dplyr::select(
      probe_id,
      parent_feature_name = feature_name,
      parent_score = score,
      parent_group_label = group_label,
      parent_adj_p = adj_p,
      selected = dplyr::all_of(flag_col)
    )

  synchronization_qc <- signature |>
    dplyr::left_join(parent_lookup, by = "probe_id")

  stopifnot(
    nrow(synchronization_qc) == nrow(signature),
    all(!is.na(synchronization_qc$parent_feature_name)),
    all(
      synchronization_qc$feature_name ==
        synchronization_qc$parent_feature_name
    ),
    isTRUE(all.equal(
      synchronization_qc$score,
      synchronization_qc$parent_score,
      tolerance = 0
    )),
    all(
      as.character(synchronization_qc$group_label) ==
        as.character(synchronization_qc$parent_group_label)
    ),
    all(synchronization_qc$selected %in% TRUE),
    all(!is.na(synchronization_qc$parent_adj_p)),
    all(is.finite(synchronization_qc$parent_adj_p)),
    all(synchronization_qc$parent_adj_p < 0.05)
  )
}

validate_metab_tables(sig_aging, difexp_aging, "aging.fdr.05")
validate_metab_tables(sig_el, difexp_el, "el.fdr.05")
validate_metab_tables(sig_survival, difexp_survival, "surv.fdr.05")
```


```{r}
oSig_aging <- OmicSignature::OmicSignature$new(
  metadata = meta_aging,
  signature = sig_aging,
  difexp = difexp_aging
)

oSig_el <- OmicSignature::OmicSignature$new(
  metadata = meta_el,
  signature = sig_el,
  difexp = difexp_el
)

oSig_survival <- OmicSignature::OmicSignature$new(
  metadata = meta_survival,
  signature = sig_survival,
  difexp = difexp_survival
)

# -----
# This is a post-construction check. The existing validate_metab_tables() checks the tables before object construction; this additional function confirms that OmicSignature did not alter or damage the probe IDs during construction. The convention requires nonmissing, character-valued, unique probe IDs.

validate_created_omic_signature <- function(object, object_name) {
  stopifnot(
    !is.null(object$signature),
    is.data.frame(object$signature),
    "probe_id" %in% names(object$signature),
    is.character(object$signature$probe_id),
    !anyNA(object$signature$probe_id),
    all(nzchar(object$signature$probe_id)),
    !anyDuplicated(object$signature$probe_id)
  )

  message(
    object_name,
    ": ",
    nrow(object$signature),
    " signature rows with valid unique probe IDs."
  )

  invisible(TRUE)
}

validate_created_omic_signature(
  oSig_aging,
  "LLFS longitudinal aging"
)

validate_created_omic_signature(
  oSig_el,
  "LLFS exceptional longevity"
)

validate_created_omic_signature(
  oSig_survival,
  "LLFS survival"
)

# -----
OmSC_LLFS <- OmicSignature::OmicSignatureCollection$new(
  OmicSigList = list(
    LongitudinalAgingMetabolites = oSig_aging,
    ExceptionalLongevityMetabolites = oSig_el,
    SurvivalAssociatedMetabolites = oSig_survival
  ),
  metadata = list(
    collection_name = make_collection_name("LLFS_Aging_EL_SurvivalMetabolites"),
    description = paste0(
      "Plasma metabolomics signatures from the Long Life Family Study, including longitudinal aging-associated, ",
      "exceptional longevity-associated, and survival-associated metabolite signatures. Feature names use RefMet standardized metabolite names. ",
      "Missing RefMet names were first rescued from original Compound.Name using the Metabolomics Workbench RefMet converter; rows still lacking RefMet names were excluded."
    ),
    author = AUTHOR
  )
)

# We validate the actual signature metadata, not temporary list names such as first or second.
llfs_members <- sigrepo_collection_members(OmSC_LLFS)

stopifnot(
  length(llfs_members) == 3L
)

actual_signature_names <- vapply(
  llfs_members,
  function(x) {
    as.character(x$metadata$signature_name)
  },
  character(1)
)

expected_signature_names <- c(
  as.character(oSig_aging$metadata$signature_name),
  as.character(oSig_el$metadata$signature_name),
  as.character(oSig_survival$metadata$signature_name)
)

stopifnot(
  setequal(
    actual_signature_names,
    expected_signature_names
  )
)

message(
  "LLFS collection contains the expected three OmicSignature objects."
)

llfs_count_qc <- tibble::tibble(
  item = c(
    "Source workbook rows",
    "Longitudinal aging signature",
    "Exceptional longevity signature",
    "Survival signature"
  ),
  observed_n = c(
    nrow(metab_raw0),
    nrow(oSig_aging$signature),
    nrow(oSig_el$signature),
    nrow(oSig_survival$signature)
  ),
  expected_n = c(
    408L,
    232L,
    212L,
    140L
  )
)

print(llfs_count_qc)

stopifnot(
  identical(
    as.integer(llfs_count_qc$observed_n),
    as.integer(llfs_count_qc$expected_n)
  ),
  nrow(difexp_aging) >= nrow(oSig_aging$signature),
  nrow(difexp_el) >= nrow(oSig_el$signature),
  nrow(difexp_survival) >= nrow(oSig_survival$signature)
)

message(
  "LLFS source and signature counts match the validated reference run."
)
```

# Approval status

No unresolved items remain for this approved reference workflow.

The reviewed RefMet rescue cache is stored with the example. The workflow
passed the OmicSignature API compatibility check, threshold validation,
pre-construction and post-construction validation, collection validation,
expected-count validation, and a clean end-to-end render.

## Save

```{r save, eval=FALSE}
saveRDS(
  oSig_aging,
  file = file.path(OUTDIR, paste0(oSig_aging$metadata$signature_name, "_refmetRescued_oSig.rds"))
)

saveRDS(
  oSig_el,
  file = file.path(OUTDIR, paste0(oSig_el$metadata$signature_name, "_refmetRescued_oSig.rds"))
)

saveRDS(
  oSig_survival,
  file = file.path(OUTDIR, paste0(oSig_survival$metadata$signature_name, "_refmetRescued_oSig.rds"))
)

saveRDS(
  OmSC_LLFS,
  file = file.path(OUTDIR, "Aging_Hs_BloodPlasma_LLFS_Aging_EL_SurvivalMetabolites_Sebastiani2024_refmetRescued_OmSC.rds")
)

cat("\nSignature sizes:\n")
print(c(
  Aging = nrow(oSig_aging$signature),
  EL = nrow(oSig_el$signature),
  Survival = nrow(oSig_survival$signature)
))

cat("\nSignature columns:\n")
print(colnames(oSig_aging$signature))
print(colnames(oSig_el$signature))
print(colnames(oSig_survival$signature))

cat("\nDifexp columns:\n")
print(colnames(oSig_aging$difexp))
print(colnames(oSig_el$difexp))
print(colnames(oSig_survival$difexp))

cat("\nGroup labels:\n")
print(table(oSig_aging$signature$group_label))
print(table(oSig_el$signature$group_label))
print(table(oSig_survival$signature$group_label))
```

```{r}
check_refmet_rescue_source <- function(sig_obj, source_df) {
  sig_obj$signature |>
    dplyr::left_join(
      source_df |>
        dplyr::select(
          probe_id,
          refmet_source
        ) |>
        dplyr::distinct(),
      by = "probe_id"
    ) |>
    dplyr::count(refmet_source)
}

check_refmet_rescue_source(oSig_aging, metab_raw)
check_refmet_rescue_source(oSig_el, metab_raw)
check_refmet_rescue_source(oSig_survival, metab_raw)
```
