---
title: "AhR/CYP1B1 MDA-MB-231 and SUM-149PT approved OmicSignature reference workflow"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

# Approval status

No unresolved items remain for this approved reference workflow.

The source interpretation and `limma.logFC` score direction are supported. The
exact signature rule is `adj.P.Val <= 0.01` with no fold-change cutoff. The
Ensembl release 114 mapping is reviewed and frozen. All four objects and the
four-member collection passed clean-session validation. Save and upload
behavior remains disabled by default.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r libraries}
library(dplyr)
library(purrr)
library(readr)
library(stringr)
library(tibble)
library(OmicSignature)
```

```{r locate-foundation}
find_repo_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    if (dir.exists(file.path(current, "codex_reference")) &&
        dir.exists(file.path(current, "codex_tests"))) {
      return(current)
    }
    parent <- dirname(current)
    if (identical(parent, current)) {
      stop("Could not locate the SigRepo repository root.", call. = FALSE)
    }
    current <- parent
  }
}

current_rmd <- knitr::current_input(dir = TRUE)
RMD_DIR <- normalizePath(
  if (nzchar(current_rmd)) dirname(current_rmd) else getwd(),
  winslash = "/",
  mustWork = TRUE
)
REPO_ROOT <- find_repo_root(RMD_DIR)
REFERENCE_DIR <- file.path(RMD_DIR, "reference_data")

source(file.path(
  REPO_ROOT, "codex_reference", "helpers", "omicsignature_compat.R"
))
sigrepo_assert_compatible_api()
```

```{r paths-and-constants}
SCENARIO_DIR <- file.path(
  REPO_ROOT,
  "codex_tests", "scenario1_diff_table", "8_AhR_CYP1B1_MDA_Sum149"
)

env_or_default <- function(name, default) {
  value <- Sys.getenv(name, unset = "")
  if (nzchar(value)) value else default
}

SOURCE_PATHS <- c(
  mda_ahr = env_or_default(
    "AHR_CYP_MDA_AHR_SOURCE",
    file.path(SCENARIO_DIR, "diffanal.MDA.AhR.xls")
  ),
  mda_cyp1b1 = env_or_default(
    "AHR_CYP_MDA_CYP1B1_SOURCE",
    file.path(SCENARIO_DIR, "diffanal.MDA.CYP.xls")
  ),
  sum149_ahr = env_or_default(
    "AHR_CYP_SUM149_AHR_SOURCE",
    file.path(SCENARIO_DIR, "diffanal.SUM.AhR.xls")
  ),
  sum149_cyp1b1 = env_or_default(
    "AHR_CYP_SUM149_CYP1B1_SOURCE",
    file.path(SCENARIO_DIR, "diffanal.SUM.CYP.xls")
  )
)

REVIEWED_MAPPING_FILE <- file.path(
  REFERENCE_DIR,
  "AhR_CYP1B1_Ensembl_v114_reviewed_mapping.rds"
)

SIGNATURE_ADJ_P_CUTOFF <- 0.01
REQUIRED_SOURCE_ROWS <- 22981L
REQUIRED_COLUMNS <- c(
  "ensembl_gene_id", "entrezgene", "hgnc_symbol", "high.class", "t",
  "P.Value", "adj.P.Val", "limma.logFC"
)

SPECS <- tribble(
  ~signature_key, ~source_file, ~cell_line, ~source_condition, ~positive_label, ~sample_type, ~signature_name, ~expected_source_signature_rows, ~expected_missing_entrez,
  "mda_ahr", "diffanal.MDA.AhR.xls", "MDA-MB-231", "AhR", "AHR_KD", "MDA-MB-231 cell", "BreastCancer_Hs_MDA_MB_231_AHR_KD_vs_CN_MontiLab2016", 219L, 3372L,
  "mda_cyp1b1", "diffanal.MDA.CYP.xls", "MDA-MB-231", "CYP1B1", "CYP1B1_KD", "MDA-MB-231 cell", "BreastCancer_Hs_MDA_MB_231_CYP1B1_KD_vs_CN_MontiLab2016", 6L, 3372L,
  "sum149_ahr", "diffanal.SUM.AhR.xls", "SUM-149PT", "AhR", "AHR_KD", "SUM-149PT cell", "BreastCancer_Hs_SUM149PT_AHR_KD_vs_CN_MontiLab2016", 2558L, 3372L,
  "sum149_cyp1b1", "diffanal.SUM.CYP.xls", "SUM-149PT", "CYP1B1", "CYP1B1_KD", "SUM-149PT cell", "BreastCancer_Hs_SUM149PT_CYP1B1_KD_vs_CN_MontiLab2016", 1077L, 3372L
)

COLLECTION_NAME <- "BreastCancer_Hs_CellLine_AhR_CYP1B1_KD_MontiLab2016"

EXPECTED_MAPPED_COUNTS <- tribble(
  ~signature_key, ~mapped_parent_rows, ~mapped_signature_rows,
  "mda_ahr", 22841L, 217L,
  "mda_cyp1b1", 22841L, 6L,
  "sum149_ahr", 22841L, 2552L,
  "sum149_cyp1b1", 22841L, 1074L
)

stopifnot(
  length(SOURCE_PATHS) == 4L,
  all(file.exists(SOURCE_PATHS)),
  nrow(SPECS) == 4L
)

sigrepo_assert_platform("transcriptomics by array")
walk(unique(SPECS$sample_type), sigrepo_assert_sample_type)
```

## Accepted preflight

The four inputs are complete limma differential-analysis tables. The accepted
contrast evidence is `cond = AhR` or `cond = CYP1B1` versus `control = CN`.
The signed score is `limma.logFC`: positive means higher expression in the
knockdown condition and negative means higher expression in `CN`. The exact
signature rule is `adj.P.Val <= 0.01`, with no fold-change cutoff.

```{r read-and-validate-sources}
clean_text <- function(x) {
  x <- as.character(x)
  x <- str_replace_all(x, "[\\u00A0\\u2007\\u202F\\uFEFF]", " ")
  str_squish(x)
}

clean_ensembl <- function(x) str_remove(clean_text(x), "\\.[0-9]+$")

read_source <- function(signature_key) {
  spec <- SPECS %>% filter(.data$signature_key == .env$signature_key)
  path <- unname(SOURCE_PATHS[[signature_key]])
  source_file <- basename(path)

  raw <- read_tsv(
    path,
    col_types = cols(.default = col_character()),
    na = c("", "NA", "N/A", "NULL"),
    progress = FALSE
  ) %>% as_tibble()

  missing_columns <- setdiff(REQUIRED_COLUMNS, names(raw))
  if (length(missing_columns) > 0L) {
    stop(source_file, " is missing required columns: ",
         paste(missing_columns, collapse = ", "), call. = FALSE)
  }
  if (nrow(raw) != REQUIRED_SOURCE_ROWS) {
    stop(source_file, " has ", nrow(raw), " rows; expected ",
         REQUIRED_SOURCE_ROWS, ".", call. = FALSE)
  }

  out <- raw %>%
    mutate(
      signature_key = signature_key,
      source_file = source_file,
      source_row = row_number() + 1L,
      original_ensembl_gene_id = clean_text(ensembl_gene_id),
      cleaned_ensembl_gene_id = clean_ensembl(ensembl_gene_id),
      source_hgnc_symbol = clean_text(hgnc_symbol),
      source_entrezgene = clean_text(entrezgene),
      score = suppressWarnings(as.numeric(limma.logFC)),
      p_value = suppressWarnings(as.numeric(P.Value)),
      adj_p = suppressWarnings(as.numeric(adj.P.Val)),
      t_statistic = suppressWarnings(as.numeric(t)),
      group_label = factor(
        case_when(
          score > 0 ~ spec$positive_label,
          score < 0 ~ "CN",
          TRUE ~ NA_character_
        ),
        levels = c(spec$positive_label, "CN")
      )
    )

  numeric_columns <- c("score", "p_value", "adj_p", "t_statistic")
  if (any(!map_lgl(out[numeric_columns], ~ all(is.finite(.x))))) {
    stop(source_file, " has missing or non-finite analytical values.",
         call. = FALSE)
  }
  if (any(out$score == 0) || anyNA(out$group_label)) {
    stop(source_file, " has zero scores or unresolved group labels.",
         call. = FALSE)
  }

  primary_identifier_missing <- (
    is.na(out$original_ensembl_gene_id) |
      !nzchar(out$original_ensembl_gene_id) |
      is.na(out$cleaned_ensembl_gene_id) |
      !nzchar(out$cleaned_ensembl_gene_id)
  )
  if (any(primary_identifier_missing)) {
    stop(source_file, " has missing primary Ensembl identifiers.",
         call. = FALSE)
  }

  missing_entrez <- sum(
    is.na(out$source_entrezgene) |
      !nzchar(out$source_entrezgene)
  )
  if (missing_entrez != spec$expected_missing_entrez) {
    stop(
      source_file, " has ", missing_entrez,
      " missing Entrez rescue identifiers; expected ",
      spec$expected_missing_entrez, ".", call. = FALSE
    )
  }

  observed_signature_rows <- sum(out$adj_p <= SIGNATURE_ADJ_P_CUTOFF)
  if (observed_signature_rows != spec$expected_source_signature_rows) {
    stop(
      source_file, " has ", observed_signature_rows,
      " source rows satisfying adj.P.Val <= 0.01; expected ",
      spec$expected_source_signature_rows, ".", call. = FALSE
    )
  }
  out
}

source_tables <- set_names(SPECS$signature_key) %>% map(read_source)
```

```{r reviewed-frozen-mapping}
if (!file.exists(REVIEWED_MAPPING_FILE)) {
  stop(
    "Reviewed frozen mapping cache is absent: ", REVIEWED_MAPPING_FILE,
    ". Run and review the one-time builder, then run the offline finalizer. ",
    "This Rmd never calls a live mapping service.",
    call. = FALSE
  )
}

mapping_bundle <- readRDS(REVIEWED_MAPPING_FILE)
required_bundle_names <- c(
  "status", "mapping", "audit", "summary", "expected_counts",
  "ensembl_release", "source_md5"
)
if (!is.list(mapping_bundle) ||
    !all(required_bundle_names %in% names(mapping_bundle))) {
  stop("Reviewed mapping cache has an invalid bundle structure.", call. = FALSE)
}
if (!identical(mapping_bundle$status, "REVIEWED")) {
  stop("Mapping cache is not marked REVIEWED.", call. = FALSE)
}
if (!identical(as.integer(mapping_bundle$ensembl_release), 114L)) {
  stop("Reviewed mapping cache is not Ensembl release 114.", call. = FALSE)
}

current_source_md5 <- unname(tools::md5sum(SOURCE_PATHS))
names(current_source_md5) <- basename(SOURCE_PATHS)
expected_source_md5 <- mapping_bundle$source_md5[names(current_source_md5)]

if (anyNA(expected_source_md5) ||
    !identical(unname(current_source_md5), unname(expected_source_md5))) {
  stop(
    "One or more source files differ from those used to build the reviewed mapping.",
    call. = FALSE
  )
}

reviewed_mapping <- as_tibble(mapping_bundle$mapping)
reviewed_audit <- as_tibble(mapping_bundle$audit)
frozen_counts <- as_tibble(mapping_bundle$expected_counts)

required_mapping_columns <- c(
  "source_file", "source_row", "original_ensembl_gene_id", "hgnc_symbol",
  "entrezgene", "cleaned_key", "target_ensembl_gene_id", "mapping_route",
  "ensembl_release", "match_status", "multiplicity", "ambiguity",
  "decision", "decision_note"
)
if (!all(required_mapping_columns %in% names(reviewed_mapping))) {
  stop("Reviewed mapping cache is missing required mapping columns.",
       call. = FALSE)
}
if (anyNA(reviewed_mapping$target_ensembl_gene_id) ||
    any(!str_detect(reviewed_mapping$target_ensembl_gene_id,
                    "^ENSG[0-9]+$"))) {
  stop("Reviewed mapping contains missing or invalid target Ensembl IDs.",
       call. = FALSE)
}
if (any(reviewed_mapping$decision != "include")) {
  stop("Reviewed mapping contains non-include rows in its mapping component.",
       call. = FALSE)
}
if (!"approval_blocking" %in% names(reviewed_audit) ||
    anyNA(reviewed_audit$approval_blocking) ||
    any(reviewed_audit$approval_blocking)) {
  stop("Reviewed mapping audit still contains approval-blocking cases.",
       call. = FALSE)
}

required_count_columns <- c(
  "signature_key", "mapped_parent_rows", "mapped_signature_rows"
)
if (!all(required_count_columns %in% names(frozen_counts)) ||
    nrow(frozen_counts) != 4L ||
    anyNA(frozen_counts[required_count_columns])) {
  stop(
    "Mapping-derived expected counts are not frozen in the reviewed cache.",
    call. = FALSE
  )
}

count_check <- frozen_counts %>%
  select(all_of(required_count_columns)) %>%
  arrange(signature_key)
expected_count_check <- EXPECTED_MAPPED_COUNTS %>%
  arrange(signature_key)
if (!identical(count_check, expected_count_check)) {
  stop(
    "Reviewed mapping counts do not match the approved frozen counts.",
    call. = FALSE
  )
}
```

```{r mapped-standardized-parents}
make_parent <- function(signature_key) {
  source <- source_tables[[signature_key]]
  spec <- SPECS %>% filter(.data$signature_key == .env$signature_key)
  map_for_source <- reviewed_mapping %>%
    filter(.data$source_file == spec$source_file)

  parent_prepared <- source %>%
    inner_join(
      map_for_source,
      by = c(
        "source_file", "source_row", "original_ensembl_gene_id",
        "source_hgnc_symbol" = "hgnc_symbol",
        "source_entrezgene" = "entrezgene"
      ),
      relationship = "many-to-many"
    ) %>%
    transmute(
      feature_name = target_ensembl_gene_id,
      symbol = source_hgnc_symbol,
      score,
      p_value,
      adj_p,
      `P.Value` = p_value,
      `adj.P.Val` = adj_p,
      t = t_statistic,
      group_label,
      source_file,
      source_row,
      original_ensembl_gene_id,
      hgnc_symbol = source_hgnc_symbol,
      entrezgene = source_entrezgene,
      cleaned_key,
      mapping_route,
      match_status,
      multiplicity,
      ambiguity,
      decision,
      decision_note
    ) %>%
    arrange(source_row, feature_name)

  expected <- frozen_counts %>%
    filter(.data$signature_key == .env$signature_key)
  if (nrow(expected) != 1L ||
      nrow(parent_prepared) != expected$mapped_parent_rows) {
    stop(signature_key, ": mapped parent count does not match frozen count.",
         call. = FALSE)
  }

  # No genuine assay identifier exists. Assign deterministic feature_n IDs
  # once on the complete mapped parent, then standardize the parent once.
  parent <- sigrepo_standardize_signature(
    as.data.frame(parent_prepared),
    paste(spec$signature_name, "parent difexp")
  )

  expected_probe_ids <- paste0("feature_", seq_len(nrow(parent)))
  if (!setequal(parent$probe_id, expected_probe_ids)) {
    stop(signature_key, ": deterministic feature_n probe IDs are invalid.",
         call. = FALSE)
  }
  parent
}

difexp_tables <- set_names(SPECS$signature_key) %>% map(make_parent)

signature_tables <- imap(difexp_tables, function(parent, signature_key) {
  expected <- frozen_counts %>%
    filter(.data$signature_key == .env$signature_key)
  child <- parent[parent$adj.P.Val <= SIGNATURE_ADJ_P_CUTOFF, , drop = FALSE]
  if (nrow(child) != expected$mapped_signature_rows) {
    stop(signature_key, ": mapped signature count does not match frozen count.",
         call. = FALSE)
  }
  child
})
```

```{r parent-child-validation}
SYNC_COLUMNS <- c(
  "probe_id", "feature_name", "score", "P.Value", "adj.P.Val", "t",
  "source_file", "source_row", "original_ensembl_gene_id", "hgnc_symbol",
  "entrezgene"
)

row_keys <- function(x, columns) {
  do.call(paste, c(lapply(x[columns], as.character), sep = "\r"))
}

iwalk(signature_tables, function(signature, signature_key) {
  parent <- difexp_tables[[signature_key]]
  spec <- SPECS %>% filter(.data$signature_key == .env$signature_key)

  stopifnot(
    all(SYNC_COLUMNS %in% names(parent)),
    all(SYNC_COLUMNS %in% names(signature)),
    all(signature$adj.P.Val <= SIGNATURE_ADJ_P_CUTOFF),
    !anyNA(parent$probe_id),
    !anyDuplicated(parent$probe_id),
    is.factor(parent$group_label),
    identical(levels(parent$group_label), c(spec$positive_label, "CN")),
    all(row_keys(signature, SYNC_COLUMNS) %in%
          row_keys(parent, SYNC_COLUMNS))
  )
})
```

```{r metadata-and-objects}
make_metadata <- function(signature_key) {
  spec <- SPECS %>% filter(.data$signature_key == .env$signature_key)
  OmicSignature::createMetadata(
    signature_name = spec$signature_name,
    organism = "Homo sapiens",
    direction_type = "bi-directional",
    assay_type = "transcriptomics",
    phenotype = paste(spec$positive_label, "versus CN in", spec$cell_line),
    platform = "transcriptomics by array",
    sample_type = spec$sample_type,
    covariates = "none",
    author = "Stefano Monti",
    year = 2016,
    adj_p_cutoff = SIGNATURE_ADJ_P_CUTOFF,
    keywords = sigrepo_keywords(c(
      "AHR", "CYP1B1", "knockdown", "breast cancer", spec$cell_line
    )),
    cutoff_description = paste0(
      "Primary repository signature criterion: adj.P.Val <= 0.01; ",
      "no fold-change cutoff."
    ),
    description = paste(
      "Unpublished MontiLab Affymetrix Human Gene 2.0 ST transcriptomic",
      "signature for", spec$positive_label, "versus CN in", spec$cell_line,
      "cells. Positive limma.logFC indicates higher expression in the",
      "knockdown condition; negative values indicate higher expression in CN."
    )
  )
}

metadata_list <- set_names(SPECS$signature_key) %>% map(make_metadata)

MEMBER_NAMES <- c(
  "MDA_MB_231_AHR_KD_vs_CN",
  "MDA_MB_231_CYP1B1_KD_vs_CN",
  "SUM149PT_AHR_KD_vs_CN",
  "SUM149PT_CYP1B1_KD_vs_CN"
)

member_key_map <- setNames(
  SPECS$signature_key,
  MEMBER_NAMES
)

osig_list <- map(
  member_key_map,
  function(signature_key) {
    OmicSignature::OmicSignature$new(
      metadata = metadata_list[[signature_key]],
      signature = signature_tables[[signature_key]],
      difexp = difexp_tables[[signature_key]],
      print_message = FALSE
    )
  }
)

collection_metadata <- list(
  collection_name = COLLECTION_NAME,
  description = paste(
    "Four unpublished MontiLab Affymetrix transcriptomic signatures for AHR",
    "and CYP1B1 knockdown versus CN in MDA-MB-231 and SUM-149PT cells."
  ),
  author = "Stefano Monti",
  year = 2016
)

ahr_cyp_collection <- OmicSignature::OmicSignatureCollection$new(
  OmicSigList = osig_list,
  metadata = collection_metadata,
  print_message = FALSE
)
```

```{r object-and-collection-validation}
stopifnot(length(osig_list) == 4L)

iwalk(osig_list, function(osig, member_name) {
  signature_key <- unname(member_key_map[[member_name]])
  expected <- frozen_counts %>%
    filter(.data$signature_key == .env$signature_key)

  stopifnot(
    length(signature_key) == 1L,
    nrow(expected) == 1L,
    nrow(osig$difexp) == expected$mapped_parent_rows,
    nrow(osig$signature) == expected$mapped_signature_rows,
    all(row_keys(osig$signature, SYNC_COLUMNS) %in%
          row_keys(osig$difexp, SYNC_COLUMNS))
  )
})

collection_members <- sigrepo_collection_members(ahr_cyp_collection)

actual_signature_names <- vapply(
  collection_members,
  function(object) {
    as.character(object$metadata$signature_name)
  },
  character(1)
)

expected_signature_names <- vapply(
  osig_list,
  function(object) {
    as.character(object$metadata$signature_name)
  },
  character(1)
)

stopifnot(
  length(collection_members) == 4L,
  setequal(actual_signature_names, expected_signature_names)
)
```

```{r save-approved-objects, eval=FALSE}
# Disabled by default for the approved reference workflow. Enable only for an
# explicitly authorized repository operation after output-path review.
stop("Saving and upload are disabled by default.", call. = FALSE)
```
